---
title: DeLorean analysis of Guo et al. mouse embryo single cell data
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Guo et al. mouse embryo data}
-->


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Guo-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")

```

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
library(rmarkdown)
load_all('..')
render('DeLorean-Guo.Rmd')

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))

```


```{r themes, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
mrc.colors <- c(
    rgb(138, 121, 103, maxColorValue=255),
    rgb(217, 165, 41 , maxColorValue=255),
    rgb(153, 152, 40 , maxColorValue=255),
    rgb(117, 139, 121, maxColorValue=255),
    rgb(33 , 103, 126, maxColorValue=255),
    rgb(208, 114, 50 , maxColorValue=255),
    rgb(106, 59 , 119, maxColorValue=255),
    rgb(130, 47 , 90 , maxColorValue=255)
)
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```



`r citet(bib[["guo_resolution_2010"]])` performed qPCR on mouse embryonic
cells at various stages.


# Data

Guo et al.'s data is available in the `DeLorean` R package.
```{r loadLib}
library(DeLorean)
data(GuoDeLorean)

```


This will load a 48 (genes) by 442 (cells) matrix containing the expression
measurements.
```{r exprDim}
dim(guo.expr)
```
and two data frames of meta data for the genes and cells.
```{r meta}
summary(guo.gene.meta)
summary(guo.cell.meta)
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes}
dl <- de.lorean(
    guo.expr,
    guo.gene.meta,
    guo.cell.meta)
dl <- estimate.hyper(
    dl,
    sigma.tau=0.5,
    delta=.5)
# knit.report(dl, "hyper-parameters")

```


# Choose genes and cells

Choose a few cells from each stage.
```{r filterCells}
set.seed(1)
num.at.each.stage <- 2
te.sampled.cells <- (
    dl$cell.meta
    %>% filter(capture < "32C" | "TE" == cell.type)
    %>% group_by(capture)
    %>% do(sample_n(., num.at.each.stage))
)
pe.sampled.cells <- (
    dl$cell.meta
    %>% filter(capture < "32C" | "PE" == cell.type | "ICM" == cell.type)
    %>% group_by(capture)
    %>% do(sample_n(., num.at.each.stage))
)
epi.sampled.cells <- (
    dl$cell.meta
    %>% filter(capture < "32C" | "EPI" == cell.type | "ICM" == cell.type)
    %>% group_by(capture)
    %>% do(sample_n(., num.at.each.stage))
)
run.model <- function(dl, cells.sampled) {
    cell.filter <- function(cells) cells %in% cells.sampled
    dl <- filter.cells(dl, cell.filter)
    dl <- format.for.stan(dl)
    dl <- compile.model.simple(dl)
    dl <- find.best.tau(dl)
    system.time(dl <- fit.model(dl))
    dl <- examine.convergence(dl)
    dl <- process.posterior(dl)
    dl <- analyse.noise.levels(dl)
    dl
}
dl.te  <- run.model(dl,  te.sampled.cells$cell)
dl.pe  <- run.model(dl,  pe.sampled.cells$cell)
dl.epi <- run.model(dl, epi.sampled.cells$cell)

```


```{r compareProfiles}
load_all('..')
dl.te  <- make.predictions(dl.te)
dl.pe  <- make.predictions(dl.pe)
dl.epi <- make.predictions(dl.epi)
gp <- plot(dl.te, type="best.predictions", profile.color='orange', add.data=F)
gp <- plot.add.profiles(gp, dl=dl.pe , color='green', genes=dl.te$genes.high.psi)
gp <- plot.add.profiles(gp, dl=dl.epi, color='blue' , genes=dl.te$genes.high.psi)
print(gp)

```

R version and packages used:
```{r Rversion}
sessionInfo()
```
