---
title: DeLorean analysis of Windram et al. Arabidopsis time series
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Arabidopsis time series}
-->

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('..')
library(rmarkdown)
render('DeLorean-Windram.Rmd')

```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Windram-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```


`r citet(bib[["windram_arabidopsis_2012"]])` assayed leaves at 24 time points
in 2 conditions.


# Data

Windram et al.'s data is available in the `DeLorean` R package.
```{r loadLib}
library(DeLorean)
data(WindramDeLorean)

```


## Obfuscate time points

Reduce resolution of observed capture time points.
```{r}
windram.cell.meta$obstime.orig <- windram.cell.meta$obstime
windram.cell.meta$capture.orig <- windram.cell.meta$capture
windram.cell.meta$obstime <- floor((windram.cell.meta$obstime-1) / 8) * 8 + 4
windram.cell.meta$capture <- (
    factor(as.character(windram.cell.meta$obstime),
           ordered=TRUE,
           levels=unique(as.character(windram.cell.meta$obstime))))

```

Just consider the *Botrytis* cells.
```{r}
dl <- de.lorean(
    windram.expr,
    windram.gene.meta,
    windram.cell.meta)
botrytis.cells <- filter(dl$cell.meta, condition == "Botrytis")$cell
cell.filter <- function(cells) cells %in% botrytis.cells
dl <- filter.cells(dl, cell.filter)

```

```{r child=report.file("data")}
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes}
dl <- estimate.hyper(
    dl,
    sigma.tau=4,
    delta=.5)

```

```{r child=report.file("hyper-parameters")}
```


# Choose genes and cells

Choose a few genes.
```{r filterCells}
set.seed(1)
num.genes <- 150
sampled.genes <- sample_n(dl$gene.meta, num.genes)$gene
gene.filter <- function(genes) genes %in% sampled.genes
dl <- filter.genes(dl, gene.filter)

```


Format the data for Stan and fit the model.
```{r fitModel}
dl <- format.for.stan(dl)
dl <- compile.model.simple(dl)
dl <- find.best.tau(dl)
system.time(dl <- fit.model(dl))
```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)

```

```{r child=report.file("convergence")}
```



# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```

```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```

```{r child=report.file("profiles")}
```


# Examine pseudotime

Did `DeLorean` learn the obfuscated pseudotime?
```{r pseudotime}
gp <- with(dl, {
    (ggplot(samples.l$tau %>% filter(iter == best.sample),
                    aes(x=tau, y=obstime.orig, color=capture),
                    environment=environment())
        + geom_point()
        + scale_x_continuous(name="Pseudotime (tau)")
        + scale_y_continuous(name="Unobfuscated time")
    )
})
png('pseudotime-vs-obfuscated.png')
print(gp)
dev.off()

```

The correlation between the obfuscated capture time with the pseudotime of
the best sample.
```{r}
with(dl$samples.l$tau %>% filter(iter == dl$best.sample),
     cor(tau, obstime.orig))

```

```{r echo=FALSE}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Windram.rds")
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
