---
title: DeLorean analysis of Trapnell et al. myoblast data
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Trapnell et al. myoblast data}
-->

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('..')
library(rmarkdown)
render('DeLorean-Trapnell.Rmd')

```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Trapnell-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))

```


Load data.
```{r loadData}
library(DeLorean)
data(TrapnellDeLorean)
dl <- de.lorean(
    trapnell.expr,
    trapnell.gene.meta,
    trapnell.cell.meta)

```

```{r child=report.file("data")}
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empBayes}
dl <- estimate.hyper(
    dl,
    sigma.tau=6,
    delta=.5)

```


```{r child=report.file("hyper-parameters")}
```


# Choose genes and cells

Select cells at random if we have too many.
```{r chooseCells}
set.seed(1)
max.cells <- 147
cells.sampled <- sample(trapnell.cell.meta$cell, max.cells)
cell.filter <- function(cells) cells %in% cells.sampled
dl <- filter.cells(dl, cell.filter)

```


Select genes by variance.
```{r chooseGenes}
max.genes <- 151
genes.high.var <- (dl$gene.var
                   %>% arrange(-psi.bar)
                   %>% head(max.genes))$gene
gene.filter <- function(genes) genes %in% genes.high.var
dl <- filter.genes(dl, gene.filter)

```


# Compile and fit model

Define and compile the model, find the best initialisation, and fit the model.
```{r model}
dl <- format.for.stan(dl)
dl <- compile.model.simple(dl)
dl <- find.best.tau(dl)
system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)

```

```{r child=report.file("convergence")}
```



# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```

```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```

```{r child=report.file("profiles")}
```


```{r echo=FALSE}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Trapnell.rds")
```


R version and packages used:
```{r sessionInfo}
sessionInfo()

```

