---
title: DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('..')
library(rmarkdown)
fit.model <- TRUE
render('Shalek-DeLorean.Rmd')

```

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data}
-->


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Shalek-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
#
# Widths for saving figures
#
text.width <- 4.79  # LaTeX width in inches
golden.ratio <- 1.618  # Pleasing ratio
fig.width <- text.width
fig.height <- text.width / golden.ratio

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```

`r date()`


`r citet(bib[["shalek_single-cell_2014"]])`


# Data

Shalek et al.'s data is available in the `DeLorean` R package.
```{r loadLib, eval=fit.model}
library(DeLorean)
data(ShalekDeLorean)

```


Build the `DeLorean` object.
```{r buildDL, eval=fit.model}
dl <- de.lorean(
    shalek.A.expr,
    shalek.A.gene.meta,
    shalek.A.cell.meta)

```


Filter out the cells we want for the time course.
```{r filterTimeCourse, eval=fit.model}
time.course.cells <- (
    dl$cell.meta
    %>% filter(! is.na(total),
               "" == assay,
               "LPS" == stimulant | "" == stimulant,
               "" == ko,
               FALSE == disrupted,
               total > 1e6,
               "" == replicate))
dl <- filter.cells(dl, function(cells) cells %in% time.course.cells$cell)

```


Only use induced genes that have been assigned to a cluster.
```{r useInducedGenes, eval=fit.model}
induced.genes <- dl$gene.meta %>% filter(! is.na(cluster))
dl <- filter.genes(dl, function(genes) genes %in% induced.genes$gene)

```
```{r child=report.file("data")}
```




# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
dl <- estimate.hyper(
    dl,
    sigma.tau=.5,
    delta=.5)

```
```{r child=report.file("hyper-parameters")}
```


# Choose genes and cells

Choose genes: take those with highest variance between time points
relative to the noise level.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Shalek.max.genes", 49))
genes.for.stan <- (
    dl$gene.var %>% arrange(-psi.bar) %>% head(max.genes))
dl <- filter.genes(dl, function(genes) genes %in% genes.for.stan$gene)

```

Choose a few cells.
```{r filterCells, eval=fit.model}
set.seed(1)
dl <- filter.cells(dl, number=getOption("Shalek.max.cells", 51))

```


Save expression data and meta data.
```{r saveInput, eval=fit.model}
saveRDS(list(expr = dl$expr, cell.meta = dl$cell.map, gene.meta=dl$gene.map),
        file='Shalek-input.rds')

```

Format the data for Stan and fit the model.
```{r fitModel, eval=fit.model}
dl <- format.for.stan(dl)
dl <- compile.model.simple(dl)
dl <- find.best.tau(dl)
system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence")}
```


# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```



```{r echo=FALSE}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Shalek.rds")
```


```{r date}
date()
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
