---
title: DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('..')
library(rmarkdown)
fit.model <- TRUE
render('Shalek-DeLorean.Rmd')

```

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data}
-->


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Shalek-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
#
# Widths for saving figures
#
text.width <- 4.79  # LaTeX width in inches
golden.ratio <- 1.618  # Pleasing ratio
fig.width <- text.width
fig.height <- text.width / golden.ratio

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```

`r date()`


`r citet(bib[["shalek_single-cell_2014"]])`


# Data

Shalek et al.'s data is available in the `DeLorean` R package.
```{r loadLib, eval=fit.model}
library(DeLorean)
data(ShalekDeLorean)

```


Build the `DeLorean` object.
```{r buildDL, eval=fit.model}
dl <- de.lorean(
    shalek.A.expr,
    shalek.A.gene.meta,
    shalek.A.cell.meta)

```


Filter out the cells we want for the time course.
```{r filterTimeCourse, eval=fit.model}
time.course.cells <- (
    dl$cell.meta
    %>% filter(! is.na(total),
               "" == assay,
               "LPS" == stimulant | "" == stimulant,
               "" == ko,
               FALSE == disrupted,
               total > 1e6,
               "" == replicate))
dl <- filter.cells(dl, function(cells) cells %in% time.course.cells$cell)

```


Only use induced genes that have been assigned to a cluster.
```{r useInducedGenes, eval=fit.model}
induced.genes <- dl$gene.meta %>% filter(! is.na(cluster))
dl <- filter.genes(dl, function(genes) genes %in% induced.genes$gene)

```
```{r child=report.file("data")}
```




# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
dl <- estimate.hyper(
    dl,
    sigma.tau=1.5,
    delta=.5,
    length.scale=5,
    model.name='simplest-model')

```
```{r child=report.file("hyper-parameters")}
```


# Choose genes and cells

Choose genes: take those with highest variance between time points
relative to the noise level.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Shalek.max.genes", 59))
core.antiviral.genes <- dl$gene.meta %>% filter("Id" == cluster)
genes.for.stan <- (
    dl$gene.var
    %>% filter(gene %in% core.antiviral.genes$gene)
    %>% arrange(-psi.bar)
    %>% head(max.genes))
dl <- filter.genes(dl, function(genes) genes %in% genes.for.stan$gene)

```

Choose a few cells but make sure we have the precocious cells.
```{r filterCells, eval=fit.model}
set.seed(1)
num.cells <- getOption("Shalek.max.cells", 51)
sampled.cells <- sample(colnames(dl$expr), num.cells)
if (! "LPS_1h_S51" %in% sampled.cells) {
    sampled.cells[1] <- "LPS_1h_S51"
}
if (! "LPS_1h_S52" %in% sampled.cells) {
    sampled.cells[2] <- "LPS_1h_S52"
}
dl <- filter.cells(dl, function(cells) cells %in% sampled.cells)

```


Save expression data and meta data.
```{r saveInput, eval=fit.model}
saveRDS(list(expr = dl$expr, cell.meta = dl$cell.map, gene.meta=dl$gene.map),
        file='Shalek-input.rds')

```

Format the data for Stan and fit the model.
```{r fitModel, eval=fit.model}
dl <- format.for.stan(dl)
dl <- compile.model(dl)
# dl <- find.best.tau(dl)
system.time(dl <- find.smooth.tau(dl))

set.seed(1)
which.tau <- 4
test.dl <- find.smooth.tau(dl, method="metropolis")
times <- data.frame(c=1:test.dl$stan.data$C,
                    tau=test.dl$tau.inits[[which.tau]]$tau)
expr.l <- with(test.dl,
    expr
    %>% melt(varnames=c("gene", "cell"), value.name="x")
    %>% mutate(gene=factor(gene, levels=levels(gene.meta$gene)),
               cell=factor(cell, levels=levels(cell.meta$cell)))
    %>% left_join(test.dl$cell.map %>% select(cell, c, capture, obstime))
    %>% left_join(times)
)
# qplot(x=times$tau, y=dl$stan.data$time)
gp <- (ggplot(expr.l, aes(x=tau, y=x, color=capture))
    + geom_point()
    + geom_line(aes(color=NULL), alpha=.5)
    + facet_wrap(~ gene))
pdf('smoothed-tau.pdf', width=10)
print(gp)
dev.off()

system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence")}
```


# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```



Save DeLorean object.
```{r echo=FALSE}
# Remove large fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Shalek.rds")
# dl <- readRDS("Shalek.rds")
```


```{r clusterAnalysis}
clustered <- dl$gene.meta %>% filter(! is.na(cluster))
clustered.expr <- shalek.A.expr[clustered$gene,time.course.cells$cell]
# rownames(clustered.expr)
# colnames(clustered.expr)
melt.expr <- function(dl, expr=dl$expr) {
    (
        expr
        %>% melt(varnames=c("gene", "cell"), value.name="x")
        %>% mutate(gene=factor(gene, levels=levels(dl$gene.meta[["gene"]])),
                   cell=factor(cell, levels=levels(dl$cell.meta[["cell"]])))
    )
}
clustered.expr.l <- (
    melt.expr(dl, clustered.expr)
    %>% left_join(dl$gene.meta))
names(clustered.expr.l)
# sample_n(clustered.expr.l, 14)
module.scores <- (
    clustered.expr.l
    %>% group_by(cluster, cell)
    %>% summarise(module.score=mean(x))
    %>% left_join(dl$samples.l$tau
                  %>% filter(dl$best.sample == iter)
                  %>% select(cell, tau)))
# Find the precocious cells
core.antiviral <- (
    module.scores
    %>% left_join(dl$cell.meta %>% select(cell, capture))
    %>% filter("Id" == cluster)
    %>% arrange(-module.score))
(ggplot(core.antiviral, aes(x=module.score, color=capture)) + geom_density())
precocious <- core.antiviral %>% filter("1h" == capture) %>% head(2)
precocious
precocious$cell
# Plot the core antiviral, the maturity, the peaked inflammation and
# the sustained inflammation module scores against pseudotime.
(ggplot(module.scores
        %>% filter(! is.na(tau),
                   cluster %in% c("Id", "IIIb", "IIIc", "IIId")),
        aes(x=tau, y=module.score, colour=cluster))
    + stat_smooth()
    + geom_point())
# Same with just core antiviral coloured by capture
(ggplot(module.scores %>% left_join(dl$cell.meta %>% select(cell, capture))
        %>% filter(! is.na(tau),
                   cluster == "Id"),
        aes(x=tau, y=module.score, colour=capture))
    + stat_smooth(aes(group=""))
    + geom_point())

# Examine what pseudotimes the model estimated for the precocious genes
( dl$samples.l$tau
    %>% filter(dl$best.sample == iter,
               cell %in% c("LPS_1h_S51", "LPS_1h_S52")))

```

```{r date}
date()
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
