---
title: DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
    # css: ../inst/Rmd/foghorn.css
---

```{r build, echo=FALSE, eval=FALSE}
#
# Execute this block to render the Rmarkdown.
#
library(devtools)
load_all('..')
library(rmarkdown)
# fit.model <- FALSE
fit.model <- TRUE
render('Shalek-DeLorean.Rmd')

```

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data}
-->


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
library(functional)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Shalek-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
#
# Widths for saving figures
#
text.width <- 4.79  # LaTeX width in inches
golden.ratio <- 1.618  # Pleasing ratio
fig.width <- text.width
fig.height <- text.width / golden.ratio

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```

`r date()`



# Data

Shalek et al.'s data `r citet(bib[["shalek_single-cell_2014"]])`
is available in the `DeLorean` R package.
```{r loadLib, eval=fit.model}
library(DeLorean)
data(ShalekDeLorean)

```


Build the `DeLorean` object.
```{r buildDL, eval=fit.model}
dl <- de.lorean(
    shalek.A.expr,
    shalek.A.gene.meta,
    shalek.A.cell.meta)

```


Filter out the cells we want for the time course.
```{r filterTimeCourse, eval=fit.model}
time.course.cells <- (
    dl$cell.meta
    %>% filter(! is.na(total),
               "" == assay,
               "LPS" == stimulant | "" == stimulant,
               "" == ko,
               FALSE == disrupted,
               total > 1e6,
               "" == replicate))
dl <- filter.cells(dl, cells=time.course.cells$cell)

```


Re-level the cells by their capture time. This improves the ordering in later
plots.
```{r relevelCells}
dl$cell.meta$cell <- factor(
    dl$cell.meta$cell,
    levels=(shalek.A.cell.meta %>% arrange(capture))$cell)

```


## Adjust by cell sizes

```{r testCellSize}
dl <- adjust.by.cell.sizes(dl)
ggplot(dl$cell.sizes, aes(x=capture, y=size)) + geom_boxplot()

```


## Choose genes and cells

Only use induced genes that have been assigned to a cluster.
```{r useInducedGenes, eval=fit.model}
induced.genes <- dl$gene.meta %>% filter(! is.na(cluster))
dl <- filter.genes(dl, genes=induced.genes$gene)

```
```{r child=report.file("data")}
```


Choose genes: take those with highest variance between time points
relative to the noise level.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Shalek.max.genes", 13))
shalek.key.genes <- unique(toupper(c(
    #
    # Cluster I d (core antiviral module; enriched for annotated antiviral and
    #             interferon response genes; for example,-
    "Ifit1", "Irf7",
    #
    # Cluster III c (peaked inflammatory module; showing rapid,
    # yet transient, induction under LPS; for example,
    "Tnf", "Il1a", "Cxcl2",
    #
    # Cluster III d (sustained inflammatory module; exhibiting
    # continued rise in expression under LPS; for example,
    "Mmp14", "Marco", "Il6",
    #
    # Cluster III b (‘maturity’ module; containing markers of
    # dendritic cell maturation; for example,
    "Cd83", "Ccr7", "Ccl22",
    #
    # At 2 h following LPS,
    "Ifnb1",
    # was bimodally expressed
    #
    # Genes encoding key inflammatory cytokines (for example,
    "Tnf", "Cxcl1",
    #
    # Figure 4: core antiviral targets.
    "Rsad2", "Stat2"
)))
clusters <- c("Id", "IIIb", "IIIc", "IIId")
# clusters <- c("Id")
dl <- analyse.variance(dl, adjust.cell.sizes=TRUE)
gene.variances <- dl$gene.var  # Save copy for later
genes.for.stan <- (
    dl$gene.var
    %>% left_join(dl$gene.meta)
    %>% mutate(key=gene %in% shalek.key.genes)
    %>% filter(cluster %in% clusters)
    %>% arrange(- Psi / Omega)
    %>% head(max.genes))
dl <- filter.genes(dl, genes=genes.for.stan$gene)
# How many come from each cluster?
qplot(genes.for.stan$cluster)

```

Choose a few cells but make sure we have the precocious cells.
```{r filterCells, eval=fit.model}
set.seed(1)
max.cells <- getOption("Shalek.max.cells", 12)
sampled.cells <- sample(colnames(dl$expr), max.cells)
if (! "LPS_1h_S51" %in% sampled.cells) {
    sampled.cells[1] <- "LPS_1h_S51"
}
if (! "LPS_1h_S52" %in% sampled.cells) {
    sampled.cells[2] <- "LPS_1h_S52"
}
dl <- filter.cells(dl, cells=sampled.cells)

```




# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
dl <- estimate.hyper(
    dl,
    sigma.tau=getOption("Shalek.sigma.tau", 1),
    delta=getOption("Shalek.delta", .5),
    length.scale=getOption("Shalek.length.scale", 5),
    model.name=getOption("Shalek.model", 'simplest-model'))

```
```{r child=report.file("hyper-parameters")}
```


Save expression data and meta data.
```{r saveInput, eval=fit.model}
saveRDS(list(expr = dl$expr, cell.meta = dl$cell.map, gene.meta=dl$gene.map),
        file='Shalek-input.rds')

```


# Fit model

Format the data for Stan and fit the model.
```{r fitModel, eval=fit.model}
dl <- format.for.stan(dl)
system.time(dl <- find.smooth.tau(dl, num.tau.to.try=detectCores()*2))
# system.time(dl <- find.smooth.tau(dl, num.cores=3, num.tau.to.try=8))
# system.time(dl <- find.smooth.tau(dl, num.tau.to.try=45))
dl <- compile.model(dl)

# library(devtools)
# load_all('..')
# test.dl <- readRDS(file='test-dl.rds')
# # orderings <- test.mh(test.dl, use.parallel=F)
# orderings <- test.mh(test.dl)
# orderings.list <- lapply(orderings, function(o) o$chain)

# set.seed(1)
# which.tau <- 4
# test.dl <- find.smooth.tau(dl, method="metropolis")
# saveRDS(test.dl, file='test-dl.rds')
# times <- data.frame(c=1:test.dl$stan.data$C,
                    # tau=test.dl$tau.inits[[which.tau]]$tau)
# expr.l <- with(test.dl,
    # expr
    # %>% melt(varnames=c("gene", "cell"), value.name="x")
    # %>% mutate(gene=factor(gene, levels=levels(gene.meta$gene)),
               # cell=factor(cell, levels=levels(cell.meta$cell)))
    # %>% left_join(test.dl$cell.map %>% dplyr::select(cell, c, capture, obstime))
    # %>% left_join(times)
# )
# # qplot(x=times$tau, y=dl$stan.data$time)
# gp <- (ggplot(expr.l, aes(x=tau, y=x, color=capture))
    # + geom_point()
    # + geom_line(aes(color=NULL), alpha=.5)
    # + facet_wrap(~ gene))
# pdf('smoothed-tau.pdf', width=10)
# print(gp)
# dev.off()

system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence")}
```


# Analyse posterior

Examine posterior and optimise best sample.
```{r posterior}
dl <- process.posterior(dl)
# dl <- optimise.best.sample(dl)
dl <- analyse.noise.levels(dl)

```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```



Save DeLorean object.
```{r echo=FALSE}
# Remove large fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Shalek.rds")
# dl <- readRDS("Shalek.rds")
```


# Cluster analysis

```{r clusterAnalysis}
clustered <- dl$gene.meta %>% filter(! is.na(cluster))
clustered.expr <- shalek.A.expr[clustered$gene,time.course.cells$cell]
# rownames(clustered.expr)
# colnames(clustered.expr)
melt.expr <- function(dl, expr=dl$expr) {
    (
        expr
        %>% melt(varnames=c("gene", "cell"), value.name="x")
        %>% mutate(gene=factor(gene, levels=levels(dl$gene.meta[["gene"]])),
                   cell=factor(cell, levels=levels(dl$cell.meta[["cell"]])))
    )
}
clustered.expr.l <- (
    melt.expr(dl, clustered.expr)
    %>% left_join(dl$gene.meta))
names(clustered.expr.l)
# sample_n(clustered.expr.l, 14)
module.scores <- (
    clustered.expr.l
    %>% group_by(cluster, cell)
    %>% summarise(module.score=mean(x))
    %>% left_join(dl$samples.l$tau
                  %>% filter(dl$best.sample == iter)
                  %>% dplyr::select(cell, tau)))
# Find the precocious cells
core.antiviral <- (
    module.scores
    %>% left_join(dl$cell.meta %>% dplyr::select(cell, capture))
    %>% filter("Id" == cluster)
    %>% arrange(-module.score))
precocious <- core.antiviral %>% filter("1h" == capture) %>% head(2)
precocious
precocious$cell
module.scores <- module.scores %>% mutate(precocious=cell %in% precocious$cell)
(ggplot(core.antiviral, aes(x=module.score, color=capture)) + geom_density())
# Plot the core antiviral, the maturity, the peaked inflammation and
# the sustained inflammation module scores against pseudotime.
(ggplot(module.scores
        %>% filter(! is.na(tau),
                   cluster %in% c("Id", "IIIb", "IIIc", "IIId")),
        aes(x=tau, y=module.score, colour=cluster))
    + stat_smooth()
    + geom_point())
# Same with just core antiviral coloured by capture
(ggplot(module.scores
        %>% left_join(dl$cell.meta %>% dplyr::select(cell, capture))
        %>% filter(! is.na(tau),
                   cluster == "Id"),
        aes(x=tau, y=module.score, colour=capture, shape=precocious))
    + stat_smooth(aes(group=""))
    + geom_point())

# Examine what pseudotimes the model estimated for the precocious genes
( dl$samples.l$tau
    %>% filter(dl$best.sample == iter,
               cell %in% precocious$cell))

```


## Check precocious

```{r checkPrecocious}
S51.dists <- with(
    dl,
    melt(expr - expr[,"LPS_1h_S51"], varnames=c("gene", "cell"))
    %>% mutate(gene=factor(gene, levels=levels(gene.map$gene)))
    %>% mutate(cell=factor(cell, levels=levels(cell.map$cell)))
    %>% left_join(gene.map)
    %>% left_join(cell.map))
names(S51.dists)
# bad.genes <- c(51, 53, 73)
bad.genes <- c()
dl$gene.map[bad.genes,]
gp <- (
    ggplot(S51.dists %>% filter(! g %in% bad.genes),
           aes(x=g, y=c, fill=value))
    + geom_tile()
    + scale_fill_gradient2()
)
png('S51-dists.png', width=960, height=960)
print(gp)
dev.off()

```


# Evaluate held out genes

Evaluate the held out genes that weren't used to fit the model.
```{r heldOutGenes}
held.out.genes <- with(dl, gene.variances
                           %>% left_join(gene.meta)
                           %>% filter(! gene %in% gene.map$gene)
                           %>% filter(cluster %in% clusters)
                           # %>% filter(cluster != "Id")
                           %>% arrange(-Psi/Omega)
                           %>% head(getOption('Shalek.held.out', 100))
)
# Get an expression matrix of held out genes in the cells of interest
held.out.expr <- shalek.A.expr[as.character(held.out.genes$gene),
                               as.character(dl$cell.map$cell)]

```


## Roughness permutation test

Permutation test for roughness.
```{r}
dl <- roughness.test(dl, held.out.expr)
print(dl$roughness.test)
# Plot the roughnesses
gp <- plot(dl, type="roughnesses")
print(gp)
ggsave('Shalek-roughnesses.png', gp)

```


```{r checkSampleVar, echo=FALSE, eval=FALSE}
# Check calculations are correct for expected sample variance from a
# Gaussian
library(matrixcalc)
library(MASS)
S <- toeplitz((10:1)/10)
set.seed(11)
R <- rWishart(1, 20, S)
Sigma <- R[,,1]
is.positive.definite(Sigma)
x <- mvrnorm(n=1e6, mu=rep(0, 10), Sigma=Sigma)
dim(x)
x.centre <- t(scale(t(x), scale=F, center=T))
vars <- rowSums(x.centre**2)/10
fit.gamma <- fitdistr(vars, 'gamma')
fit.lognormal <- fitdistr(vars, 'lognormal')
(
    ggplot(data.frame(V=vars), aes(x=V))
    + geom_density()
    + stat_function(fun=Curry(dgamma,
                              shape=fit.gamma$estimate['shape'],
                              rate=fit.gamma$estimate['rate']),
                    linetype='dashed')
    + stat_function(fun=Curry(dlnorm,
                              meanlog=fit.lognormal$estimate['meanlog'],
                              sdlog=fit.lognormal$estimate['sdlog']),
                    linetype='dotted')
)
mean(log(vars))
sd(log(vars))
mean(diag(Sigma)) - mean(as.vector(Sigma))

```

```{r tauPosteriorOffset.2, echo=FALSE}
gp <- (
    ggplot(dl$samples.l$tau,
            aes(x=tau.offset, color=capture))
    + geom_density()
    + geom_rug(alpha=.01)
    + stat_function(fun=Curry(dnorm, sd=dl$stan.data$sigma_tau),
                    linetype=2,
                    color="black")
)
ggsave('Shalek-tau-offset.png', gp)

```


```{r totalVar, echo=FALSE, eval=FALSE}
expr.l <- (
    dl$expr
    %>% melt(varnames=c("gene", "cell"), value.name="x")
    %>% left_join(dl$cell.meta))
expr.gene.time <- (
    expr.l
    %>% group_by(gene, capture)
    %>% summarise(num=n(), m=mean(x), v=mean((x-m)**2)))
# Must use weighted means
expr.gene <- (
    expr.gene.time
    %>% group_by(gene)
    %>% summarise(m.m=weighted.mean(m, num),
                  Psi=weighted.mean((m-m.m)**2, num),
                  Omega=weighted.mean(v, num)))
(
    expr.l
    %>% group_by(gene)
    %>% summarise(v.total=mean((x-mean(x))**2))
    %>% left_join(expr.gene)
    %>% mutate(v.total-Psi-Omega))

```


```{r date}
date()
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
