---
title: DeLorean analysis of Tang et al. ESC data
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of ESC derivation}
-->

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('..')
library(rmarkdown)
render('Tang-DeLorean.Rmd')

```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Tang-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
#
# Widths for saving figures
#
text.width <- 4.79  # LaTeX width in inches
golden.ratio <- 1.618  # Pleasing ratio
fig.width <- text.width
fig.height <- text.width / golden.ratio

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```


`r citet(bib[["tang_tracing_2010"]])` assayed leaves at 24 time points
in 2 conditions.


# Data

Tang et al.'s data is available in the `DeLorean` R package.
```{r loadLib}
library(DeLorean)
data(TangDeLorean)
dl <- de.lorean(
    tang.rna.seq,
    tang.rna.seq.gene.meta,
    tang.rna.seq.cell.meta)

```
```{r child=report.file("data")}
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes}
dl <- estimate.hyper(
    dl,
    sigma.tau=1.5,
    delta=.5)

```
```{r child=report.file("hyper-parameters")}
```


## Choose cells and genes

Select some cells at random if we have too many.
```{r sampleCells}
set.seed(1)
max.at.each.stage <- min(getOption("Tang.max.at.stage", nrow(dl$cell.meta)))
sample.up.to <- function(.data, size) {
    if (size < nrow(.data)) {
        sample_n(.data, size)
    } else {
        .data
    }
}
sampled.cells <- (
    dl$cell.meta
    %>% group_by(capture)
    %>% do(sample.up.to(., max.at.each.stage)))
sampled.cells

dl <- filter.cells(dl, function(cells) cells %in% sampled.cells$cell)

```

Select some genes at random if we have too many.
```{r sampleGenes}
max.genes <- min(getOption("Tang.max.genes", nrow(dl$gene.meta)))
if (max.genes <= length(tang.key.genes)) {
    sampled.genes <- (dl$gene.meta
                      %>% filter(key)
                      %>% sample_n(max.genes))$gene
} else {
    sampled.genes <- sample(dl$gene.meta$gene, max.genes)
}
dl <- filter.genes(dl, function(genes) genes %in% sampled.genes)

```


Save expression data and meta data.
```{r}
saveRDS(list(expr = dl$expr, cell.meta = dl$cell.map, gene.meta=dl$gene.map),
        file='Tang-input.rds')

```


Format the data for Stan and fit the model.
```{r fitModel}
dl <- format.for.stan(dl)
dl <- compile.model.simple(dl)
dl <- find.best.tau(dl)
system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)

```
```{r child=report.file("convergence")}
```



# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```


```{r echo=FALSE}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Tang.rds")
# dl <- readRDS("Windram.rds")

```


```{r date}
date()
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
