---
title: DeLorean analysis of Windram et al. Arabidopsis time series
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
    # css: ../inst/Rmd/foghorn.css
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Arabidopsis time series}
-->

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('..')
library(rmarkdown)
# fit.model <- FALSE
fit.model <- TRUE
render('Windram-DeLorean.Rmd')

```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Windram-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
#
# Widths for saving figures
#
text.width <- 4.7  # LaTeX width in inches
golden.ratio <- 1.618  # Pleasing ratio
fig.width <- text.width
fig.height <- text.width / golden.ratio
html5 <- list(width=1300, height=700)  # in pixels
html5$ratio <- with(html5, width / height)
slide.fig.width <- 7

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```

`r date()`

`r citet(bib[["windram_arabidopsis_2012"]])` assayed leaves at 24 time points
in 2 conditions.


# Data

Windram et al.'s data is available in the `DeLorean` R package.
```{r loadLib}
library(DeLorean)
data(WindramDeLorean)

```


## Obfuscate time points

Reduce resolution of observed capture time points.
```{r eval=fit.model}
group.size <- 12
windram.cell.meta$obstime.orig <- windram.cell.meta$obstime
windram.cell.meta$capture.orig <- windram.cell.meta$capture
windram.cell.meta$obstime <- (
    floor((windram.cell.meta$obstime-1) / group.size) * group.size
    + group.size / 2)
windram.cell.meta$capture <- (
    factor(as.character(windram.cell.meta$obstime),
           ordered=TRUE,
           levels=unique(as.character(windram.cell.meta$obstime))))

```

Just consider the *Botrytis* cells and select some at random if we have
too many.
```{r eval=fit.model}
dl <- de.lorean(
    windram.expr,
    windram.gene.meta,
    windram.cell.meta)
botrytis.cells <- dl$cell.meta %>% filter(condition == "Botrytis")
dl <- filter.cells(dl, cells=botrytis.cells$cell)

```


```{r sampleCells, eval=fit.model}
set.seed(getOption("Windram.seed", 1))
max.cells <- min(getOption("Windram.max.cells", 24))
dl <- filter.cells(dl, number=max.cells)

```

```{r child=report.file("data")}
```


## Adjust by cell sizes

```{r testCellSize}
dl <- adjust.by.cell.sizes(dl)
ggplot(dl$cell.sizes, aes(x=capture, y=size)) + geom_boxplot()

```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
dl <- estimate.hyper(
    dl,
    sigma.tau=group.size / 2,
    delta=.5,
    model.name='simplest-model')
fits.omega <- test.fit(dl$gene.var$Omega)
print(fits.omega$gp)
fits.psi <- test.fit(dl$gene.var$Psi)
print(fits.psi$gp)

```

```{r child=report.file("hyper-parameters")}
```


# Choose genes

Choose a few genes.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Windram.max.genes", 100))
sampled.genes <- sample_n(dl$gene.meta, max.genes)$gene
gene.filter <- function(genes) genes %in% sampled.genes
dl <- filter.genes(dl, gene.filter)

```


Format the data for Stan and fit the model.
```{r fitModel, eval=fit.model}
dl <- format.for.stan(dl)
# Save for input to Monocle
saveRDS(list(expr = dl$expr, cell.meta = dl$cell.map, gene.meta=dl$gene.map),
        file='Windram-input.rds')
dl <- compile.model(dl)
# dl <- find.smooth.tau(dl)
dl <- find.best.tau(dl)
system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)

```

```{r child=report.file("convergence")}
```



# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```


# Examine pseudotime

Did `DeLorean` learn the obfuscated pseudotime?
```{r pseudotime}
gp.pseudo <- with(dl, {
    (ggplot(samples.l$tau %>% filter(iter == best.sample),
                    aes(x=tau, y=obstime.orig, color=capture),
                    environment=environment())
        + geom_point()
        # + scale_x_continuous(name="pseudotime")
        # + scale_y_continuous(name="capture time")
        # + scale_color_discrete(name="low-resolution\ngroup")
        + scale_y_continuous(name="true capture time")
        + scale_x_continuous(name="pseudotime")
        + scale_color_discrete(name="model\ncapture\ntime")
    )
})
print(gp.pseudo)
# png('pseudotime-vs-obfuscated.png')
ggsave('Windram-pseudotime-vs-obfuscated.pdf', gp.pseudo,
       width=text.width, height=text.width,
       dpi=300, units="in")
ggsave('Windram-pseudotime-vs-obfuscated.png', gp.pseudo,
       width=slide.fig.width, height=slide.fig.width / html5$ratio,
       dpi=300, units="in")
# Save convergence as well
ggsave('Windram-Rhat.png', plot(dl, type="convergence"),
       width=slide.fig.width, height=slide.fig.width / html5$ratio,
       dpi=300, units="in")
# Save profiles as well
gp.profiles <- plot(dl, type="profiles")
ggsave('Windram-profiles.pdf', gp.profiles,
       width=text.width * 2, height=text.width * 2,
       dpi=150, units="in")
ggsave('Windram-profiles.png', gp.profiles,
       width=slide.fig.width * 2, height=slide.fig.width * 2/ html5$ratio,
       dpi=150, units="in")
# Save tau posterior
gp.tau.posterior <- (
    ggplot(dl$samples.l$tau, aes(x=capture.orig, y=tau, color=capture))
    + geom_boxplot()
    # + theme(axis.text.x = element_text(angle = 90, hjust = 1))
    # + theme_gray(base_size=24)
    + scale_x_discrete(name="true capture time")
    + scale_y_continuous(name="pseudotime")
    + scale_color_discrete(name="model\ncapture\ntime")
    + coord_flip())
ggsave('Windram-tau-posterior.pdf', gp.tau.posterior,
       width=text.width, height=text.width,
       dpi=300, units="in")
ggsave('Windram-tau-posterior.png', gp.tau.posterior,
       width=slide.fig.width, height=slide.fig.width / html5$ratio,
       dpi=300, units="in")
# Save best tau
gp.tau.best <- plot(dl, type="pseudotime")
ggsave('Windram-tau-best.pdf', gp.tau.best,
       width=text.width, height=text.width,
       dpi=300, units="in")
ggsave('Windram-tau-best.png', gp.tau.best,
       width=slide.fig.width, height=slide.fig.width / html5$ratio,
       dpi=300, units="in")

```


Look at the expected correlation between the obfuscated capture time with the
pseudotime in the full posterior.
```{r}
posterior.cor <- (
    dl$samples.l$tau
    %>% group_by(iter)
    %>% summarise(pseudotime.capture.cor=cor(tau, obstime.orig,
                                             method="spearman"))
)
posterior.cor.mean <- mean(posterior.cor$pseudotime.capture.cor)
posterior.cor.best <- filter(posterior.cor,
                             dl$best.sample == iter)$pseudotime.capture.cor
gp.post.cor <- (ggplot(posterior.cor, aes(x=pseudotime.capture.cor))
    + geom_histogram()
    + geom_vline(x=.94, linetype='dashed', color='red')
    + scale_x_continuous(name="correlation"))
print(gp.post.cor)
ggsave('Windram-posterior-cor.pdf', plot=gp.post.cor,
       width=fig.width, height=fig.height)
ggsave('Windram-posterior-cor.png', gp.post.cor,
       width=slide.fig.width / 2, height=slide.fig.width / 2 / html5$ratio,
       dpi=300, units="in")

```


The correlation between the obfuscated capture time with the pseudotime of
the best sample.
```{r}
with(dl$samples.l$tau %>% filter(iter == dl$best.sample),
     cor(tau, obstime.orig, method="spearman"))

```

# Evaluate held out genes

Fit some held out genes.
```{r heldOutGenes}
held.out.genes <- with(dl, gene.var
                           %>% left_join(gene.meta)
                           %>% filter(! gene %in% gene.map$gene)
                           %>% arrange(-Psi/Omega)
                           %>% head(getOption('Windram.held.out', 100)))
held.out.expr <- windram.expr[as.character(held.out.genes$gene),
                              as.character(dl$cell.map$cell)]

```


## Roughness permutation test

Permutation test for roughness.
```{r}
dl <- roughness.test(dl, held.out.expr)
print(dl$roughness.test)
# Plot the roughnesses
gp <- plot(dl, type="roughnesses")
print(gp)
ggsave('Windram-roughnesses.png', gp)

```


```{r echo=FALSE}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Windram.rds")
# dl <- readRDS("Windram.rds")

```

```{r savePlots, echo=FALSE, eval=FALSE}
plots <- with(dl, list(
    tau.posterior = (
        ggplot(samples.l$tau %>% arrange(capture),
            aes(x=cell, y=tau, color=capture))
        + geom_boxplot()
        + coord_flip()
    ),
    profiles = plot(dl, type="profiles")
))
saveRDS(plots, "Windram-plots.rds")

```

R version and packages used:
```{r Rversion}
date()
sessionInfo()
```
