```{r hyperAttach, include=FALSE}
# Attach the DeLorean data frame to access members
attach(dl)
```

## Fixed hyperparameters

The standard deviation of cell pseudotimes around the cell capture times has been set as
$$
    \sigma_\tau = `r opts$sigma.tau`
$$
The length scale of the gene expression profiles has been set as
$$
l = `r hyper$l`
$$


## Variance decomposition into temporal variation and noise

We examine the variation of expression both between- and within-time points. We
expect some of the variation of expression within a time point to be caused by
variation in cells' pseudotimes and genes' mean profiles. The rest of the
variation within a time point is assumed to be noise.  Conversely we expect
most of the variation between time points to be due variation in genes' mean
profiles although some will be due to noise.

Let
$$
k_c \in \{\kappa_1, \dots, \kappa_T\}
$$
be the time point at which cell $c$ was captured. We partition the cells
by their captured time points indexed by $1 \le t \le T$:
$$
\mathcal{K}_t = \{c: k_c = \kappa_t\}
$$
We group the expression measurements by gene and observed capture time to
calculate means and variances:
$$
\begin{align}
\mathbb{M}_{g,t} &= \text{Mean}_{c \in \mathcal{K}_t}\{x_{g,c}\} \\\\
\mathbb{V}_{g,t} &= \text{Var}_{c \in \mathcal{K}_t}\{x_{g,c}\}
\end{align}
$$

### Noise estimates
We estimate the gene-specific noise levels by assuming that all the within-time
variation in the data is due to noise. This should be a slight overestimate as
some variation of expression will be due to variation in the pseudotemporal
dimension.
$$
  \hat{\omega}_g = \textrm{Mean}_t\{\mathbb{V}_{g,t}\}
$$
Giving us this fit for our empirical Bayes prior on the $\log \hat{\omega}_g$
(the blue curve is the prior on $\log \omega_g$ and the density and rug plots
show the distribution of the $\log \hat{\omega}_g$ estimated using empirical
Bayes above).
```{r plotOmegaPrior}
ggplot(
    gene.var,
    aes(x = log(omega.hat))) +
  geom_density() +
  geom_rug() +
  stat_function(
    fun = function(x) dnorm(x, mean = hyper$mu_omega, sd = hyper$sigma_omega),
    colour = "blue", alpha = .7, linetype = "dashed")
```


### Temporal variation estimates
We estimate the temporal variation by calculating the expected variance of
samples at the given capture times from the expression profile of a gene $g$. These depend on our fixed length scale and unknown temporal variance $\psi_g$.
The covariance of these samples will be $\psi_g \hat{\Sigma}$ where
$$
  \hat{\Sigma}_{t_1,t_2} = \Sigma_\tau(\kappa_{t_1}, \kappa_{t_2})
$$
The expected variance of our samples is $\psi_g V_{\hat{\Sigma}}$ where
$$
  V_{\hat{\Sigma}} = \textrm{Mean}\{\textrm{Diag}(\hat{\Sigma})\} - \textrm{Mean}\{\hat{\Sigma}\}
$$
and we can slightly overestimate the temporal variances $\psi_g$ by ignoring the
noise in our data
$$
  \hat{\psi}_g = \frac{\textrm{Var}_t\{\mathbb{M}_{g,t}\}}{V_{\hat{\Sigma}}}
$$
Giving us this fit for our empirical Bayes prior on the temporal variation
$\log \hat{\psi}_g$
```{r psiPrior}
ggplot(
    gene.var,
    aes(x = log(psi.hat))) +
  geom_density() +
  geom_rug() +
  stat_function(
    fun = function(x) dnorm(x, mean = hyper$mu_psi, sd = hyper$sigma_psi),
    colour = "blue", alpha = .7, linetype = "dashed")
```


### Correlation between temporal variation and noise

The correlation between the temporal variation and noise estimates is
`r with(gene.var, round(cor(psi.hat, omega.hat), digits = 2))`.
Plotting the estimated noise levels against temporal variation gives:
```{r geneWithinVsBetweenVar}
ggplot(gene.var, aes(x = log10(psi.hat), y = log10(omega.hat))) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = .7)
```


## Cell sizes relation to gene variation

How big are the cell size estimates relative to the average variation in the gene expression? The vertical line represents the square root of the mean
variance of the gene expressions.
```{r cellSizeRelExpr}
gene.vars <- apply(expr, 1, var)
mean.gene.variation <- sqrt(mean(gene.vars))
ggplot(cell.sizes, aes(x = S.hat)) +
    geom_histogram(bins = 30) +
    geom_vline(xintercept = mean.gene.variation)
```





```{r hyperDetach, include = FALSE}
# Detach the previously attached DeLorean data frame
detach(dl)
```
