```{r hyperAttach, include=FALSE}
# Attach the DeLorean data frame to access members
attach(dl)
```

## Fixed hyperparameters

The standard deviation of cell pseudotimes around the cell capture times has been set as
$$
    \sigma_\tau = `r opts$sigma.tau`
$$
The length scale of the gene expression profiles has been set as
$$
l = `r hyper$l`
$$


## Variance decomposition into temporal variation and noise

We examine the variance both between- and within-time points. We expect some of
the variation within a time point to be associated with noise in the temporal
dimension, that is estimating the pseudotime correctly should reduce the
within-time variance. The rest of the variation within a time point will be due
to noise.  Conversely we expect most of the variation between time points to be
due to temporal variation although some will be due to noise.

Let
$$
k_c \in \{\kappa_1, \dots, \kappa_T\}
$$
be the time point at which cell $c$ was captured. We partition the cells
by their captured time points indexed by $1 \le t \le T$:
$$
\mathcal{K}_t = \{c: k_c = \kappa_t\}
$$
We group the expression measurements by gene and observed capture time to
calculate means and variances:
$$
\begin{align}
\mathbb{M}_{g,t} &= \text{Mean}_{c \in \mathcal{K}_t}\{x_{g,c}\} \\\\
\mathbb{V}_{g,t} &= \text{Var}_{c \in \mathcal{K}_t}\{x_{g,c}\}
\end{align}
$$

### Noise estimates

We estimate the gene-specific noise levels by assuming that all the within-time
variation in the data is due to noise. This should be a slight overestimate as
some will be due to noise in the pseudotemporal dimension.
$$
  \hat{\omega}_g = \textrm{Mean}_t\{\mathbb{V}_{g,t}\}
$$
The estimated standard deviations $\sqrt{\hat{\omega}_g}$ have this distribution:
```{r geneWithinVar}
(ggplot(gene.var, aes(x=sqrt(omega.hat))) + geom_density() + geom_rug())
```

Giving us this fit for our empirical Bayes prior on the $\log \hat{\omega}_g$
```{r omegaPrior}
(ggplot(gene.var, aes(x=log(omega.hat))) + geom_density() + geom_rug()
    + stat_function(fun=function(x) dnorm(x,
                                          mean=hyper$mu_omega,
                                          sd=hyper$sigma_omega),
                    colour="blue", alpha=.7, linetype="dashed")
)

```


### Temporal variation estimates

We estimate the temporal variation by calculating the expected variance of samples
at the capture times from the expression profile of a gene $g$. These depend on
our fixed length scale and unknown temporal variance $\psi_g$. The
covariance of these samples will be $\psi_g \hat{\Sigma}$ where
$$
  \hat{\Sigma}_{t_1,t_2} = \Sigma_\tau(\kappa_{t_1}, \kappa_{t_2})
$$
The expected variance of our samples is $\psi_g V_{\hat{\Sigma}}$ where
$$
  V_{\hat{\Sigma}} = \textrm{Mean}\{\textrm{Diag}(\hat{\Sigma})\} - \textrm{Mean}\{\hat{\Sigma}\}
$$
and we can slightly overestimate the temporal variances $\psi_g$ by ignoring the
noise in our data
$$
  \hat{\psi}_g = \frac{\textrm{Var}_t\{\mathbb{M}_{g,t}\}}{V_{\hat{\Sigma}}}
$$
The estimated standard deviations $\sqrt{\hat{\psi}_g}$ have this distribution:
```{r geneBetweenVar}
(ggplot(gene.var, aes(x=sqrt(psi.hat))) + geom_density() + geom_rug())
```

Giving us this fit for our empirical Bayes prior on the temporal variation
$\log \hat{\psi}_g$
```{r psiPrior}
(ggplot(gene.var, aes(x=log(psi.hat))) + geom_density() + geom_rug()
    + stat_function(fun=function(x) dnorm(x,
                                          mean=hyper$mu_psi,
                                          sd=hyper$sigma_psi),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

### Correlation between temporal variation and noise

The correlation between the temporal variation and noise estimates is
`r with(gene.var, cor(psi.hat, omega.hat))`.
Plotting the estimated noise levels against temporal variation gives:
```{r geneWithinVsBetweenVar}
(
    ggplot(gene.var, aes(x=sqrt(psi.hat), y=sqrt(omega.hat)))
    + geom_point()
    + geom_abline(intercept=0, slope=1, linetype="dashed", alpha=.7)
)
```

```{r hyperDetach, include=FALSE}
# Detach the previously attached DeLorean data frame
detach(dl)
```
