
# Posterior


## Priors vs. posteriors

The posteriors are compared to the priors. First the posterior for the cell-specific
drop out effects $\alpha_c$:
```{r posteriorAlpha}
(qplot(geom="density", x=as.vector(la$alpha)) + geom_rug()
    + stat_function(fun=function(x) dnorm(x,
                                          mean=hyper$mu_alpha,
                                          sd=hyper$sigma_alpha),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

The posterior for the the gene-specific drop out effects $\beta_g$:
```{r posteriorBeta}
(qplot(geom="density", x=as.vector(la$beta)) + geom_rug()
    + stat_function(fun=function(x) dnorm(x,
                                          sd=hyper$sigma_beta),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

The posterior for the cell size factors $S_c$:
```{r posteriorS}
(qplot(geom="density", x=as.vector(la$S)) + geom_rug()
    + stat_function(fun=function(x) dnorm(x,
                                          mean=hyper$mu_S,
                                          sd=hyper$sigma_S),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

The posterior for the gene means $\phi_g$:
```{r posteriorPhi}
(qplot(geom="density", x=as.vector(la$phi)) + geom_rug()
    + stat_function(fun=function(x) dnorm(x,
                                          mean=hyper$mu_phi,
                                          sd=hyper$sigma_phi),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

The posterior for the between time variances $\psi_g$:
```{r posteriorPsi}
(qplot(geom="density", x=as.vector(la$psi)) + geom_rug()
    + stat_function(fun=function(x) dlnorm(x,
                                           mean=hyper$mu_psi,
                                           sd=hyper$sigma_psi),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

The posterior for the within time variances $\omega_g$:
```{r posteriorOmega}
(qplot(geom="density", x=as.vector(la$psi)) + geom_rug()
    + stat_function(fun=function(x) dlnorm(x,
                                           mean=hyper$mu_omega,
                                           sd=hyper$sigma_omega),
                    colour="blue", alpha=.7, linetype="dashed")
)
```

## Other posteriors

The posterior for the pseudotimes $\tau_c$:
```{r posteriorTau}
qplot(geom="density", x=as.vector(la$tau)) + geom_rug()
```

The posterior for the log likelihoods of the held-out data:
```{r posteriorLL}
qplot(geom="density", x=as.vector(la$ll)) + geom_rug()
```

The posterior for the predicted means:
```{r posteriorMean}
qplot(geom="density", x=as.vector(la$predictedmean)) + geom_rug(alpha=.1)
```

The posterior for the model log probability:
```{r posteriorLP}
qplot(geom="density", x=as.vector(la$lp__)) + geom_rug()
```

The measurement noise compared to the temporal variation:
```{r noiseScatter}
(ggplot(noise.levels,
        aes(x=psi, y=omega, color=gene %in% genes.high.psi),
        environment=environment()) + geom_point())
```

Measurement noise:
```{r measurementNoise}
(ggplot(noise.levels,
        aes(x=gene, y=log(omega), fill=gene %in% genes.high.psi),
        environment=environment())
    + geom_boxplot())
```

Temporal variation:
```{r temporalVariation}
(ggplot(noise.levels,
        aes(x=gene, y=log(psi), fill=gene %in% genes.high.psi),
        environment=environment())
    + geom_boxplot())
```

Measurement noise against temporal variation (summarised by gene):
```{r measurementNoiseByGene}
(ggplot(gene.noise.levels,
        aes(x=psi, y=omega, color=gene %in% genes.high.psi),
        environment=environment())
    + geom_point())
```
