```{r posteriorAttach, include = FALSE}
# Attach the DeLorean data frame to access members
attach(dl)
```

## Posterior

### Pseudotime

The estimated pseudotime from the best sample against the observed
capture time.
```{r tauBest}
plot(dl, type = "pseudotime")
```

The full tau posterior for each cell.
```{r tauPosterior}
ggplot(
    samples.l$tau %>% arrange(capture),
    aes(x = reorder(cell, tau, FUN = median), y = tau, color = capture)) +
  geom_boxplot() +
  labs(x = 'cell') +
  coord_flip()
```

The posterior of the pseudotime offsets relative to the capture times.
We also show the prior on the offset.
```{r tauPosteriorOffset}
plot(dl, type = "tau.offsets")
```

### Between time variation

The posterior for the between time variation $\log \psi_g$:
```{r posteriorPsi}
ggplot(samples.l$psi, aes(x = log(psi))) +
  geom_density() +
  geom_rug() +
  stat_function(
    fun = function(x) dnorm(x, mean = hyper$mu_psi, sd = hyper$sigma_psi),
    colour = "blue", alpha = .7, linetype = "dashed")
```

Temporal variation by gene:
```{r temporalVariation}
ggplot(
    noise.levels,
    aes(
      x = reorder(gene, psi, FUN = median),
      y = log10(psi),
      fill = gene %in% genes.high.psi),
    environment = environment()) +
  geom_boxplot() +
  labs(x = 'gene') +
  coord_flip()
```


### Within time variation

The posterior for the within time variation $\log \omega_g$:
```{r posteriorOmega}
ggplot(samples.l$omega, aes(x = log(omega))) +
  geom_density() +
  geom_rug() +
  stat_function(
    fun = function(x) dnorm(x, mean = hyper$mu_omega, sd = hyper$sigma_omega),
    colour = "blue", alpha = .7, linetype = "dashed")
```

Within time variation by gene:
```{r measurementNoise}
ggplot(
    noise.levels,
    aes(x = reorder(gene, omega, FUN = median), y = log10(omega), fill = gene %in% genes.high.psi),
    environment = environment()) +
  geom_boxplot() +
  labs(x = 'gene') +
  coord_flip()
```


### Variation comparisons

The within time variation compared to the temporal variation (on log-log scale):
```{r noiseScatter}
ggplot(
    noise.levels,
    aes(x = log(psi), y = log(omega), color = gene %in% genes.high.psi),
    environment = environment()) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", alpha = .7)
```


```{r echo = FALSE}
have.cell.sizes <- 'S' %in% names(samples.l)
```
```{r child = report.file("cellsizes"), eval = have.cell.sizes}
```


### Other posteriors

The posterior for the model log probability:
```{r posteriorLP}
ggplot(samples.l$lp__, aes(x = lp__)) +
  geom_density() +
  geom_rug()
```


```{r posteriorDetach, include = FALSE}
# Detach the previously attached DeLorean data frame
detach(dl)
```
