---
title: DeLorean analysis of Trapnell et al. myoblast data
author: John Reid
bibliography: DeLorean.bib
output: html_document
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Trapnell et al. myoblast data}
-->

```{r build, echo=FALSE, eval=FALSE}
devtools::load_all('../..')
devtools::load_all('../../../DeLoreanData')
rmarkdown::render('DeLorean-Trapnell.Rmd')

```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Trapnell-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}

```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
library(DeLoreanData)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file(file.path('Rmd', 'foghorn.css'),
                                               package="DeLorean"))

```


Load data.
```{r loadData}
library(DeLorean)
data(TrapnellDeLorean)
dl <- de.lorean(
    trapnell.expr,
    trapnell.gene.meta,
    trapnell.cell.meta)

```

```{r child=report.file("data")}
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empBayes}
dl <- estimate.hyper(dl, sigma.tau=6)
```


```{r child=report.file("hyper-parameters")}
```


# Choose genes and cells

Select cells at random if we have too many.
```{r filterCells}
set.seed(1)
max.cells <- min(getOption("DL.max.cells", 57))
sampled.cells <- sample(dl$cell.meta$cell, max.cells)
cell.filter <- function(cells) cells %in% sampled.cells
dl <- filter.cells(dl, cell.filter)

```


Select genes by variance.
```{r chooseGenes}
max.genes <- 151
genes.high.var <- (dl$gene.var
                   %>% arrange(-Psi)
                   %>% head(max.genes))$gene
gene.filter <- function(genes) genes %in% genes.high.var
dl <- filter.genes(dl, gene.filter)

```


# Compile and fit model

Define and compile the model, find the best initialisation, and fit the model.
```{r model}
dl <- prepare.for.stan(dl)
dl <- compile.model(dl)
dl <- find.best.tau(dl)
system.time(dl <- fit.model(dl))

```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)

```

```{r child=report.file("convergence")}
```



# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)

```

```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)

```

```{r child=report.file("profiles")}
```

Plot some specific profiles mentioned in Trapnell et al.
```{r}
genes.fig.2 <- c("MEF2C", "MYH2", "CDK1", "ID1", "MYOG")
genes.fig.3.cluster <- c("HES1", "PBX1", "MYOG", "CDK1", "AKT1", "MYF6")
genes.fig.3.reg <- c("MEF2", "MYOD", "P300", "E47", "MEIS1", "PBX1",
                     "XBP1", "CUX1", "USF1", "ZIC1", "MZF1", "POU2F1",
                     "POU3F2", "AHR", "RREB1", "RFX1", "HIVEP2", "PATZ1",
                     "DDIT3", "BACHA1", "ARID5B")
genes.fig.3 <- c(genes.fig.3.cluster, genes.fig.3.reg)
genes.fig.4.a <- c("MZF1", "ZIC1", "XBP1", "CUX1", "ARID5B", "USF1", "POU2F1",
                   "AHR", "PATZ1", "DDIT3", "BACH1", "MYOG")
genes.fig.all <- unique(c(genes.fig.2, genes.fig.3, genes.fig.4.a))
filter(dl$gene.map , hgnc_symbol %in% genes.fig.all)
filter(dl$gene.meta, hgnc_symbol %in% genes.fig.all)
png('Trapnell-profiles.png', width=1200, height=800)
(plot(dl,
      type="profiles",
      genes=filter(dl$gene.map, hgnc_symbol %in% genes.fig.all)$gene)
    + facet_wrap(~ hgnc_symbol + gene))
dev.off()
```


```{r echo=FALSE}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Data/Trapnell.rds")
```


R version and packages used:
```{r sessionInfo}
sessionInfo()
```

