---
title: Oscope analysis of McDavid et al. cell cycle data
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r build, echo=FALSE, eval=FALSE}
#
# Execute this block to render the script.
#
devtools::load_all('../..')
devtools::load_all('../../../DeLoreanData')
rmarkdown::render('McDavid-Oscope.Rmd')
```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/McDavid-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
source(system.file("scripts/shared.R", package="DeLorean"))
```

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
# suppressMessages(loadfonts())
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)
```


`r citet(bib[["mcdavid_modeling_2014"]])` assayed actively cycling single
cells in order to examine the confounding effect of the cell cycle on
single cell expression analyses. They measured the expression of 333 genes
in 930 cells across three cell cycle phases and three cell lines.


## Data

McDavid et al.'s data is available in the `DeLorean` R package. Not all 333
genes are represented as several are rarely expressed.
```{r loadLib}
library(Oscope)
# vignette('Oscope_vignette')
library(DeLoreanData)
library(dplyr)
library(reshape2)
library(ggplot2)
data(McDavidDeLorean)
seed <- getOption("McDavid.seed", 1)
set.seed(seed)
gene.levels <- levels(mcdavid.gene.meta$gene)
cell.levels <- levels(mcdavid.cell.meta$cell)
```

## Choose genes and cells
Choose a few cells from the PC3 cell line and those genes with -log pvalue greater than 10.
```{r chooseCells}
cells.PC3 <- mcdavid.cell.meta %>% filter("PC3" == cellline)
genes.sig <- filter(mcdavid.gene.meta, pvalue > 10)
# Shift positively expressed genes closer to unexpressed genes to improve fit.
# Just as in DeLorean analysis
expr <- pmax(mcdavid.expr - .69, 0)[as.character(genes.sig$gene),as.character(cells.PC3$cell)]
```

## Normalisation
Oscope uses expression values on the original scale, not log transformed and expects
mean gene levels to be above 100.
```{r normalise}
oExpr <- 1000*exp(expr) # Scale to put on similar scale to Oscope expected
Sizes = MedianNorm(oExpr)
DataNorm <- GetNormalizedMat(oExpr, Sizes)
```

# Mean-variance
We perform the mean-variance analysis but we are going to use all the genes in any case,
so we don't use the results.
```{r meanVar}
MV <- CalcMV(Data=oExpr, Sizes=Sizes)
print(length(MV$GeneToUse))
```

# Paired sine model
Rescale the genes expression for the test.
```{r normPairedSine}
DataInput <- NormForSine(DataNorm)
```
Perform the paired sine test.
```{r pairedSine}
SineRes <- OscopeSine(DataInput, parallel=TRUE)
str(SineRes)
```

# K-medoids algorithm
Use all gene pairs (`quan=0`) as all the genes have small p-values in McDavid's
test of periodicity.
```{r kMedoids}
KMRes <- OscopeKM(SineRes, quan=0)
print(KMRes)
```
Check the p-values of genes identified.
```{r checkGenes}
filter(mcdavid.gene.meta, gene %in% unlist(KMRes)) %>% arrange(pvalue)
```

# Flag clusters
Flag the clusters that are not significant.
```{r flagClusters}
ToRM <- FlagCluster(SineRes, KMRes, DataInput)
print(ToRM$FlagID_bysine)
print(ToRM$FlagID_byphase)
print(ToRM$FlagID) # all flagged clusters
KMResUse <- KMRes[-ToRM$FlagID]
print(KMResUse)
```

# Extended nearest insertion
Calculate orderings for each cluster.
```{r insertion}
ENIRes <- OscopeENI(KMRes=KMResUse, Data=DataInput, NCThre=1000, parallel=TRUE)
```
Reorder the data.
```{r reorder}
DataOrder <- DataNorm[,ENIRes[["cluster1"]]]
expr.l <-
    melt(DataOrder, varnames=c('gene', 'cell'), value.name='x') %>%
    mutate(
        gene=factor(gene, levels=gene.levels),
        cell=factor(cell, levels=cell.levels)) %>%
    left_join(data.frame(cell=factor(colnames(DataOrder), levels=cell.levels),
                         time=1:ncol(DataOrder)))
sample_n(expr.l, 10)
```
Plot the genes.
```{r plotGenes}
ggplot(expr.l, aes(x=time, y=log(x))) + geom_line() + facet_wrap(~ gene)
```

## Session information
```{r date}
date()
```
R version and packages used:
```{r Rversion}
devtools::session_info()
```
