---
title: Oscope analysis of McDavid et al. cell cycle data
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r build, echo=FALSE, eval=FALSE}
#
# Execute this block to render the script.
#
devtools::load_all('../..')
devtools::load_all('../../../DeLoreanData')
rmarkdown::render('McDavid-Oscope.Rmd')
```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/McDavid-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
source(system.file("scripts/shared.R", package="DeLorean"))
```

```{r init, echo=FALSE, message=FALSE, warning=FALSE}
# suppressMessages(loadfonts())
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)
```


`r citet(bib[["mcdavid_modeling_2014"]])` assayed actively cycling single
cells in order to examine the confounding effect of the cell cycle on
single cell expression analyses. They measured the expression of 333 genes
in 930 cells across three cell cycle phases and three cell lines.


## Data

McDavid et al.'s data is available in the `DeLorean` R package. Not all 333
genes are represented as several are rarely expressed.
```{r loadLib}
library(Oscope)
library(DeLoreanData)
library(dplyr)
library(reshape2)
library(ggplot2)
data(McDavidDeLorean)
seed <- getOption("McDavid.seed", 1)
set.seed(seed)
gene.levels <- levels(mcdavid.gene.meta$gene)
cell.levels <- levels(mcdavid.cell.meta$cell)
```

## Choose genes and cells

Choose a few cells from the PC3 cell line and those genes with -log pvalue greater than 10.
```{r chooseCells}
cells.PC3 <- mcdavid.cell.meta %>% filter("PC3" == cellline)
# Shift positively expressed genes closer to unexpressed genes to improve fit.
expr <- pmax(mcdavid.expr - .69, 0)[,as.character(cells.PC3$cell)]
```

## Normalise by median values.
```{r normalise}
oExpr <- exp(expr)
Sizes = MedianNorm(oExpr)
# Sizes = QuantileNorm(oExpr, .75)
DataNorm <- GetNormalizedMat(oExpr, Sizes)

MV <- data.frame(mean=rowMeans(oExpr), var=apply(oExpr, 1, var))
ggplot(MV, aes(x=mean, y=var)) + geom_point()
# MV <- CalcMV(Data=oExpr, Sizes=Sizes, MeanCutHigh=NULL, Plot=TRUE)

expr.l <-
    expr %>%
    melt(varnames=c('gene', 'cell'), value.name='x') %>%
    mutate(gene=factor(gene, levels=gene.levels),
           cell=factor(cell, levels=cell.levels))
sizes <-
    expr.l %>%
    group_by(gene) %>%
    dplyr::summarise(prop.expr=mean(x>0)) %>%
    filter(prop.expr > .5) %>%
    left_join(expr.l) %>%
    group_by(gene) %>%
    dplyr::summarise(mu=mean(x)) %>%
    left_join(expr.l) %>%
    group_by(cell) %>%
    dplyr::summarise(size=median(x-mu))
Sizes <- sizes$size
names(Sizes) <- sizes$cell
expr.l <-
    expr.l %>%
    left_join(sizes) %>%
    mutate(x.adj=x-size)
```

## Examine mean and variance of genes
```{r meanVar}
gene.mean.var <-
    expr.l %>%
    group_by(gene) %>%
    dplyr::summarise(mu=mean(x.adj), var=var(x.adj)) %>%
    left_join(mcdavid.gene.meta)
ggplot(gene.mean.var, aes(x=mu, y=var, color=pvalue<10)) + geom_point()
```


## Choose genes
```{r chooseGenes}
genes.sig <- mcdavid.gene.meta %>% filter(pvalue > 10)
expr.l <- filter(expr.l, gene %in% genes.sig$gene)
expr <- acast(expr.l, gene ~ cell, value.var='x.adj')
dim(expr)
```

## Normalise for paired-sine
```{r normPairedSine}
DataInput <- NormForSine(DataNorm)
# DataInput <- NormForSine(DataNorm, qt1=0, qt2=1)
```

## Paired sine
```{r pairedSine}
SineRes <- OscopeSine(DataInput, parallel=TRUE)
str(SineRes)
```

## K-medoids algorithm
```{r kMedoids}
KMRes <- OscopeKM(SineRes, quan=.5)
print(KMRes)
```
Check the p-values of genes identified.
```{r checkGenes}
filter(mcdavid.gene.meta, gene %in% unlist(KMRes))
```

## Flag clusters
```{r flagClusters}
ToRM <- FlagCluster(SineRes, KMRes, DataInput)
print(ToRM$FlagID_bysine)
print(ToRM$FlagID_byphase)
print(ToRM$FlagID) # all flagged clusters
KMResUse <- KMRes[-ToRM$FlagID]
print(KMResUse)
```

## Extended nearest insertion
Calculate orderings for each cluster.
```{r insertion}
ENIRes <- OscopeENI(KMRes = KMRes, Data = DataInput, NCThre = 100, parallel=TRUE)
print(ENIRes)
```

Compare orderings for the two clusters.
```{r compareOrderings}
ordering.invert <- function(ordering) {
    result <- rep(0, length(ordering))
    for (n in 1:length(result)) {
        result[ordering[n]] <- n
    }
    result
}
orderings <- data.frame(order1=ENIRes[['cluster1']], order2=ENIRes[['cluster2']])
invert1 <- ordering.invert(orderings$order1)
invert2 <- ordering.invert(orderings$order2)
ggplot(mutate(orderings, yend=invert2[order1]),
       aes(y=1:nrow(orderings), yend=yend)) +
    geom_segment(x=0, xend=1)
cell.order <- data.frame(cell=factor(colnames(expr), levels=cell.levels),
                         pos1=(1:nrow(orderings))[invert1],
                         pos2=(1:nrow(orderings))[invert2])
```

Plot genes.
```{r plotGenes}
ggplot(expr.l %>% left_join(cell.order) %>% filter(gene %in% KMRes[['cluster1']]),
       aes(x=pos1, y=x.adj, color=gene)) +
    geom_point()

cluster1 <- expr[KMRes[['cluster1']], ENIRes[['cluster1']]]
cluster2 <- expr[KMRes[['cluster2']], ENIRes[['cluster2']]]
qplot(x=1:ncol(expr), y=cluster1[1,])
qplot(x=1:ncol(expr), y=cluster1[2,])
qplot(x=1:ncol(expr), y=cluster2[1,])
qplot(x=1:ncol(expr), y=cluster2[2,])
```


## Session information
```{r date}
date()
```
R version and packages used:
```{r Rversion}
devtools::session_info()
```
