---
title: DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r build, echo=FALSE, eval=FALSE}
#
# Execute this block to render the Rmarkdown.
#
devtools::load_all('../..')
# fit.model <- FALSE
fit.model <- TRUE
rmarkdown::render('Shalek-DeLorean.Rmd')
```

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Shalek et al. primary mouse bone-marrow-derived dendritic cells data}
-->


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
library(functional)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Shalek-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
source(system.file("scripts/shared.R", package="DeLorean"))
```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file(file.path('Rmd', 'foghorn.css'),
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)
```

`r date()`



# Data

Shalek et al.'s data `r citet(bib[["shalek_single-cell_2014"]])`
is available in the `DeLorean` R package.
```{r loadLib, eval=fit.model}
library(DeLorean)
data(ShalekDeLorean)
```


Build the `DeLorean` object.
```{r buildDL, eval=fit.model}
dl <- de.lorean(
    shalek.A.expr,
    shalek.A.gene.meta,
    shalek.A.cell.meta)
```


Filter out the cells we want for the time course.
```{r filterTimeCourse, eval=fit.model}
time.course.cells <- (
    dl$cell.meta
    %>% filter(! is.na(total),
               "" == assay,
               "LPS" == stimulant | "" == stimulant,
               "" == ko,
               FALSE == disrupted,
               total > 1e6,
               "" == replicate))
dl <- filter.cells(dl, cells=time.course.cells$cell)
```


Re-level the cells by their capture time. This improves the ordering in later
plots.
```{r relevelCells}
dl$cell.meta$cell <- factor(
    dl$cell.meta$cell,
    levels=(shalek.A.cell.meta %>% arrange(capture))$cell)
```


## Adjust by cell sizes

```{r testCellSize}
dl <- adjust.by.cell.sizes(dl)
ggplot(dl$cell.sizes, aes(x=capture, y=size)) + geom_boxplot()
```


## Choose genes and cells

Only use induced genes that have been assigned to a cluster.
```{r useInducedGenes, eval=fit.model}
induced.genes <- dl$gene.meta %>% filter(! is.na(cluster))
dl <- filter.genes(dl, genes=induced.genes$gene)
```
```{r child=report.file("data")}
```


Choose genes: take those with highest variance between time points
relative to the noise level.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Shalek.max.genes", 74))
shalek.key.genes <- unique(toupper(c(
    #
    # Cluster I d (core antiviral module; enriched for annotated antiviral and
    #             interferon response genes; for example,-
    "Ifit1", "Irf7",
    #
    # Cluster III c (peaked inflammatory module; showing rapid,
    # yet transient, induction under LPS; for example,
    "Tnf", "Il1a", "Cxcl2",
    #
    # Cluster III d (sustained inflammatory module; exhibiting
    # continued rise in expression under LPS; for example,
    "Mmp14", "Marco", "Il6",
    #
    # Cluster III b (‘maturity’ module; containing markers of
    # dendritic cell maturation; for example,
    "Cd83", "Ccr7", "Ccl22",
    #
    # At 2 h following LPS,
    "Ifnb1",
    # was bimodally expressed
    #
    # Genes encoding key inflammatory cytokines (for example,
    "Tnf", "Cxcl1",
    #
    # Figure 4: core antiviral targets.
    "Rsad2", "Stat2"
)))
clusters <- c("Id", "IIIb", "IIIc", "IIId")
# clusters <- c("Id")
dl <- analyse.variance(dl, adjust.cell.sizes=TRUE)
gene.variances <- dl$gene.var  # Save copy for later
genes.for.stan <- (
    dl$gene.var
    %>% left_join(dl$gene.meta)
    %>% mutate(key=gene %in% shalek.key.genes)
    %>% filter(cluster %in% clusters)
    %>% arrange(- Psi / Omega)
    %>% head(max.genes))
dl <- filter.genes(dl, genes=genes.for.stan$gene)
# How many come from each cluster?
qplot(genes.for.stan$cluster)
```

Choose a few cells but make sure we have the precocious cells.
```{r filterCells, eval=fit.model}
set.seed(getOption("Shalek.seed", 1))
max.cells <- getOption("Shalek.max.cells", 37)
sampled.cells <- sample(colnames(dl$expr), max.cells)
if (! "LPS_1h_S51" %in% sampled.cells) {
    sampled.cells[1] <- "LPS_1h_S51"
}
if (! "LPS_1h_S52" %in% sampled.cells) {
    sampled.cells[2] <- "LPS_1h_S52"
}
dl <- filter.cells(dl, cells=sampled.cells)
```




# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
dl <- estimate.hyper(
    dl,
    sigma.tau=getOption("Shalek.sigma.tau", 1),
    length.scale=getOption("Shalek.length.scale", 5),
    model.name=getOption("Shalek.model", 'simplest-model'))
```
```{r child=report.file("hyper-parameters")}
```


Save expression data and meta data.
```{r saveInput, eval=fit.model}
saveRDS(list(expr=dl$expr,
             cell.meta=dl$cell.map,
             gene.meta=dl$gene.map,
             gene.variances=gene.variances),
        file='Data/Shalek-input.rds')
```


# Fit model

Format the data for Stan and fit the model.
```{r fitModel, eval=fit.model}
dl <- prepare.for.stan(dl)
dl <- compile.model(dl)
system.time(dl <- find.best.tau(dl))
system.time(dl <- fit.model(dl, getOption("Shalek.method", "vb")))
```


# Examine convergence.

```{r examConv}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence")}
```


# Analyse posterior

Examine posterior and optimise best sample.
```{r posterior}
dl <- process.posterior(dl)
# dl <- optimise.best.sample(dl)
dl <- analyse.noise.levels(dl)
```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```



Save DeLorean object.
```{r echo=FALSE}
# Remove large fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Data/Shalek.rds")
# dl <- readRDS("Data/Shalek.rds")
```


# Cluster analysis

```{r clusterAnalysis}
clustered <- dl$gene.meta %>% filter(! is.na(cluster))
fitted.time.course.cells <- (
    time.course.cells
    %>% filter(cell %in% dl$cell.map[['cell']])
)$cell
clustered.expr <- shalek.A.expr[as.character(clustered$gene),
                                as.character(fitted.time.course.cells)]
# rownames(clustered.expr)
# colnames(clustered.expr)
clustered.expr.l <- melt.expr(dl, clustered.expr) %>% left_join(dl$gene.meta)
names(clustered.expr.l)
# sample_n(clustered.expr.l, 14)
module.scores <- (
    clustered.expr.l
    %>% group_by(cluster, cell)
    %>% dplyr::summarise(module.score=mean(x))
    %>% left_join(dl$samples.l$tau
                  %>% filter(dl$best.sample == iter)
                  %>% dplyr::select(cell, tau)))
module.scores
stopifnot(all(! is.na(module.scores)))
# Find the precocious cells
core.antiviral <- (
    module.scores
    %>% left_join(dl$cell.meta %>% dplyr::select(cell, capture))
    %>% filter("Id" == cluster)
    %>% arrange(-module.score))
precocious <- core.antiviral %>% filter("1h" == capture) %>% head(2)
precocious
precocious$cell
module.scores <- (
    module.scores %>% mutate(type=ifelse(cell %in% precocious$cell,
                                         "precocious",
                                         "not precocious"))
)
plots <- list()
plots$core.antiviral <- (
    ggplot(core.antiviral,
           aes(x=module.score, color=capture))
    + geom_density()
)
print(plots$core.antiviral)
ggsave('Shalek-core-antiviral.pdf', plots$core.antiviral + plos.theme,
       width=2*fig.width, height=2*fig.height)
# Plot the core antiviral, the maturity, the peaked inflammation and
# the sustained inflammation module scores against pseudotime.
plots$module <- (
    ggplot(module.scores
           %>% filter(! is.na(tau),
                      cluster %in% c("Id", "IIIb", "IIIc", "IIId")),
           aes(x=tau, y=module.score, colour=cluster))
    + stat_smooth()
    + geom_point()
)
ggsave('Shalek-module.pdf', plots$module + plos.theme,
       width=2*fig.width, height=2*fig.height)
print(plots$module)
# Same with just core antiviral coloured by capture
plots$core <- (
    ggplot(module.scores
           %>% left_join(dl$cell.meta %>% dplyr::select(cell, capture))
           %>% filter(! is.na(tau),
                      cluster == "Id"),
           aes(x=tau, y=module.score, colour=capture, shape=type))
    + stat_smooth(aes(group="", color=NULL))
    + geom_point()
)
print(plots$core)
ggsave('Shalek-core.svg', plots$core + plos.theme,
       width=plos.width, height=plos.height, units="in")
ggsave('Shalek-core.tiff', plots$core + plos.theme,
       width=plos.width, height=plos.height, units="in", dpi=300)
# Cannot use EPS as has transparency
# ggsave('Shalek-core.eps', plots$core + plos.theme,
#        width=plos.width, height=plos.height, units="in")
ggsave('Shalek-core.pdf', plots$core + plos.theme,
       width=plos.width, height=plos.height)
# Examine what pseudotimes the model estimated for the precocious genes
(
    dl$samples.l$tau
    %>% filter(dl$best.sample == iter,
               cell %in% precocious$cell)
)
```


## Check precocious

```{r checkPrecocious}
S51.dists <- with(
    dl,
    reshape2::melt(expr - expr[,"LPS_1h_S51"], varnames=c("gene", "cell"))
    %>% mutate(gene=factor(gene, levels=levels(gene.map$gene)))
    %>% mutate(cell=factor(cell, levels=levels(cell.map$cell)))
    %>% left_join(gene.map)
    %>% left_join(cell.map))
names(S51.dists)
# bad.genes <- c(51, 53, 73)
bad.genes <- c()
dl$gene.map[bad.genes,]
gp <- (
    ggplot(S51.dists %>% filter(! g %in% bad.genes),
           aes(x=g, y=c, fill=value))
    + geom_tile()
    + scale_fill_gradient2()
)
png('S51-dists.png', width=960, height=960)
print(gp)
dev.off()
print(gp)
```


# Evaluate held out genes

Evaluate the held out genes that weren't used to fit the model.
```{r heldOutGenes}
held.out.genes <- with(dl, gene.variances
                           %>% left_join(gene.meta)
                           %>% filter(! gene %in% gene.map$gene)
                           %>% filter(cluster %in% clusters)
                           # %>% filter(cluster != "Id")
                           %>% arrange(-Psi/Omega)
                           %>% head(getOption('Shalek.held.out', 100))
)
# Get an expression matrix of held out genes in the cells of interest
held.out.expr <- shalek.A.expr[as.character(held.out.genes$gene),
                               as.character(dl$cell.map$cell)]
```


## Roughness permutation test

Permutation test for roughness.
```{r}
dl <- roughness.test(dl, held.out.expr)
print(dl$roughness.test)
plots$roughnesses <- plot(dl, type="roughnesses")
print(plots$roughnesses)
ggsave('Shalek-roughnesses.svg', plots$roughnesses + plos.theme,
       width=plos.width, height=plos.height, units="in")
ggsave('Shalek-roughnesses.tiff', plots$roughnesses + plos.theme,
       width=plos.width, height=plos.height, units="in", dpi=300)
ggsave('Shalek-roughnesses.pdf', plots$roughnesses + plos.theme,
       width=plos.width, height=plos.height, units="in")
ggsave('Shalek-roughnesses.png', plots$roughnesses + plos.theme)
```


```{r checkSampleVar, echo=FALSE, eval=FALSE}
# Check calculations are correct for expected sample variance from a
# Gaussian
library(matrixcalc)
library(MASS)
S <- toeplitz((10:1)/10)
set.seed(11)
R <- rWishart(1, 20, S)
Sigma <- R[,,1]
is.positive.definite(Sigma)
x <- mvrnorm(n=1e6, mu=rep(0, 10), Sigma=Sigma)
dim(x)
x.centre <- t(scale(t(x), scale=F, center=T))
vars <- rowSums(x.centre**2)/10
fit.gamma <- fitdistr(vars, 'gamma')
fit.lognormal <- fitdistr(vars, 'lognormal')
(
    ggplot(data.frame(V=vars), aes(x=V))
    + geom_density()
    + stat_function(fun=Curry(dgamma,
                              shape=fit.gamma$estimate['shape'],
                              rate=fit.gamma$estimate['rate']),
                    linetype='dashed')
    + stat_function(fun=Curry(dlnorm,
                              meanlog=fit.lognormal$estimate['meanlog'],
                              sdlog=fit.lognormal$estimate['sdlog']),
                    linetype='dotted')
)
mean(log(vars))
sd(log(vars))
mean(diag(Sigma)) - mean(as.vector(Sigma))
```

```{r tauPosteriorOffset.2, echo=FALSE}
plots$tau.offset <- (
    ggplot(dl$samples.l$tau,
            aes(x=tau.offset, color=capture))
    + geom_density()
    + geom_rug(alpha=.01)
    + stat_function(fun=Curry(dnorm, sd=dl$stan.data$sigma_tau),
                    linetype=2,
                    color="black")
)
ggsave('Shalek-tau-offset.png', plots$tau.offset + plos.theme)
```


```{r totalVar, echo=FALSE, eval=FALSE}
expr.l <- melt.expr(dl) %>% left_join(dl$cell.meta)
expr.gene.time <- (
    expr.l
    %>% group_by(gene, capture)
    %>% dplyr::summarise(num=n(), m=mean(x), v=mean((x-m)**2)))
# Must use weighted means
expr.gene <- (
    expr.gene.time
    %>% group_by(gene)
    %>% dplyr::summarise(m.m=weighted.mean(m, num),
                  Psi=weighted.mean((m-m.m)**2, num),
                  Omega=weighted.mean(v, num)))
(
    expr.l
    %>% group_by(gene)
    %>% dplyr::summarise(v.total=mean((x-mean(x))**2))
    %>% left_join(expr.gene)
    %>% mutate(v.total-Psi-Omega))
```


Test plotting held out genes.
```{r heldOutPlot}
held.out.genes <- held.out.select.genes(dl, shalek.A.expr, 25)
held.out <- held.out.melt(dl, shalek.A.expr, held.out.genes)
system.time(posterior <- held.out.posterior(dl, held.out))
gp <- plot.held.out.posterior(dl, posterior)
print(gp)
```

Save plots.
```{r savePlots}
saveRDS(plots, "Data/Shalek-plots.rds")
```

```{r date}
date()
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
