---
title: Test tau smoothing algorithms
author: John Reid
output: html_document
---

```{r build, echo=FALSE, eval=FALSE}
library(devtools)
load_all('../..')
library(rmarkdown)
render('test-smooth-tau.Rmd')

```


```{r loadLibs, echo=FALSE, message=FALSE}
library(stringr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(DeLorean)

```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/smooth-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Widths for saving figures
#
text.width <- 4.79  # LaTeX width in inches
golden.ratio <- 1.618  # Pleasing ratio
fig.width <- text.width
fig.height <- text.width / golden.ratio
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file("inst/Rmd/foghorn.css",
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)

```



Load test data and create log likelihood ordering smoothing function.
```{r load.test.data}
library(devtools)
load_all('../..')
test.dl <- readRDS(file='test-dl.rds')
# test.dl <- readRDS(file='Shalek-input.rds')
# test.dl <- format.for.stan(test.dl)
ll.fn <- ordering.log.likelihood.fn(test.dl)

```


Try some random samples to see what log likelihoods they have.
```{r randomSamples}
random.lls <- function(num.samples=1000) {
    sapply(1:num.samples, function(i) ll.fn(sample(test.dl$stan.data$C)))
}
lls.random <- data.frame(stats="random", ll=random.lls())

```


Maximise some random samples to see what log likelihoods they have.
```{r maximiseRandomSamples}
maximise.lls <- function(num.samples=10) {
    sapply(1:num.samples,
           function(i) ll.fn(ordering.maximise(sample(test.dl$stan.data$C),
                                               ll.fn)))
}
lls.maximised <- data.frame(stats="maximised", ll=maximise.lls())

```


Try a Metropolis-Hastings run.
```{r mh-run}
init.ordering <- sample(test.dl$stan.data$C)
mh.run <- ordering.metropolis.hastings(
    init.ordering,
    ll.fn,
    proposal.fn=ordering.random.block.move,
    iterations=1000,
    thin=25)
lls.mh <- data.frame(stats="Metropolis", ll=mh.run$log.likelihoods)

```


Random projections.
```{r randomProj}
expr.center <- t(scale(t(test.dl$expr), scale=FALSE, center=TRUE))
expr.moments <- rbind(expr.center, expr.center^2)
unit.vector <- function(G) {
    #u <- runif(G)
    u <- rnorm(G)
    #u <- runif(G) - .5
    u <- u / sqrt(sum(u^2))
}
random.projection.lls <- function(suff.stats, num.projs=1000) {
    sapply(1:num.projs,
           function(i) ll.fn(order(unit.vector(nrow(suff.stats))
                                   %*% suff.stats)))
}
maximise.random.projection.lls <- function(suff.stats, num.projs=50) {
    sapply(1:num.projs,
           function(i) ll.fn(ordering.maximise(order(unit.vector(nrow(suff.stats)) %*% suff.stats), ll.fn)))
}
lls.random.projs <- rbind(
    data.frame(stats="RP.uncentred", ll=random.projection.lls(test.dl$expr)),
    data.frame(stats="RP.centred", ll=random.projection.lls(expr.center)),
    data.frame(stats="RP.maximised", ll=maximise.random.projection.lls(expr.center)),
    data.frame(stats="RP.moments", ll=random.projection.lls(expr.moments)))

```


Plot the distributions of log likelihoods of the orderings from the
different methods.
```{r plotLLs}
(ggplot(lls.random
        %>% rbind(lls.maximised)
        %>% rbind(lls.mh)
        %>% rbind(lls.random.projs)
        %>% filter("RP.maximised" != stats,
                   "maximised" != stats),
        aes(x=ll, color=stats))
    + geom_density())

```


```{r date}
date()
```


R version and packages used:
```{r Rversion}
sessionInfo()
```
