---
title: DeLorean analysis of Windram et al. Arabidopsis time series
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Arabidopsis time series}
-->

```{r build, echo=FALSE, eval=FALSE}
devtools::load_all('../..')
# fit.model <- FALSE
fit.model <- TRUE
rmarkdown::render('Windram-DeLorean.Rmd')
```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Windram-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
source(system.file("scripts/shared.R", package="DeLorean"))
```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file(file.path('Rmd', 'foghorn.css'),
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)
```

`r date()`

`r citet(bib[["windram_arabidopsis_2012"]])` assayed leaves at 24 time points
in 2 conditions.


# Data

Windram et al.'s data is available in the `DeLorean` R package.
```{r loadLib}
library(DeLorean)
library(dplyr)
library(ggplot2)
data(WindramDeLorean)
seed <- getOption("Windram.seed", 1)
set.seed(seed)
```


## Obfuscate time points

Reduce resolution of observed capture time points.
```{r eval=fit.model}
group.size <- 12
windram.cell.meta$obstime.orig <- windram.cell.meta$obstime
windram.cell.meta$capture.orig <- windram.cell.meta$capture
windram.cell.meta$obstime <- (
    floor((windram.cell.meta$obstime-1) / group.size) * group.size
    + group.size / 2)
windram.cell.meta$capture <- (
    factor(as.character(windram.cell.meta$obstime),
           ordered=TRUE,
           levels=unique(as.character(windram.cell.meta$obstime))))
```

Just consider the *Botrytis* cells
```{r eval=fit.model}
dl <- de.lorean(
    windram.expr,
    windram.gene.meta,
    windram.cell.meta)
botrytis.cells <- dl$cell.meta %>% filter(condition == "Botrytis")
dl <- filter.cells(dl, cells=botrytis.cells$cell)
```


# Estimate hyperparameters

Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
model.name <- getOption('Windram.model', 'exact')
dl <- estimate.hyper(dl, sigma.tau=group.size/4, model.name=model.name, adjust.cell.sizes=FALSE)
if (! dl$opts$model.estimates.cell.sizes) {
    dl <- adjust.by.cell.sizes(dl)
}
fits.omega <- test.fit(dl$gene.var$omega.hat)
print(fits.omega$gp)
fits.psi <- test.fit(dl$gene.var$psi.hat)
print(fits.psi$gp)
```
```{r child=report.file("data")}
```
```{r child=report.file("hyper-parameters")}
```


# Choose cells

Select some cells at random if we have too many.
```{r sampleCells, eval=fit.model}
# 24 is all the cells
max.cells <- min(getOption("Windram.max.cells", 24))
dl <- filter.cells(dl, number=max.cells)
```


# Choose genes

Choose a few (at most 100) genes to fit the model.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Windram.max.genes", 100))
dl <- filter.genes(dl, number=max.genes)
```
Choose some held out genes to validate the model with.
```{r heldOutGenes}
held.out.genes <- with(dl, gene.var
                           %>% left_join(gene.meta)
                           %>% filter(! gene %in% rownames(dl$expr))
                           %>% arrange(-psi.hat/omega.hat)
                           %>% head(getOption('Windram.held.out', 100)))
held.out.expr <- windram.expr[as.character(held.out.genes$gene),
                              colnames(dl$expr)]
```


# Fit model

```{r options}
num.inits <- getOption("Windram.num.inits", default.num.cores())
inf.method=getOption("Windram.method", "vb")
```

Define and compile the model, find the best initialisation, and fit the model.
The model name is "`r model.name`", the inference method is "`r inf.method`"
with `r num.inits` initialisations, the seed is `r seed` and the data have
`r dim(dl)[1]` genes and `r dim(dl)[2]` cells.

```{r fitModel, eval=fit.model}
dl <- prepare.for.stan(dl, num.inducing.tau = 30)
# Save for input to Monocle
saveRDS(list(expr=dl$expr,
             held.out.expr=held.out.expr,
             cell.meta=dl$cell.map,
             gene.meta=dl$gene.map),
        file='Data/Windram-input.rds')
dl <- compile.model(dl)
dl <- find.good.ordering(dl, seriation.find.orderings)
plot(dl, type='orderings')
dl <- pseudotimes.from.orderings(dl, num.to.keep=num.inits)
system.time(dl <- fit.model(dl, method=inf.method, num.inits=num.inits))
plots <- list()
if ("vb" == inf.method) {
    plots$pair.plot <- pseudotimes.pair.plot(dl)
}
```


```{r examConv, eval=("sample" == inf.method)}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence"), eval=("sample" == inf.method)}
```



# Analyse posterior

Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)
```
```{r child=report.file("posterior")}
```


# Profiles

Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```


# Examine pseudotime

Did `DeLorean` learn the obfuscated pseudotime?
```{r pseudotime}
plots$pseudo <- with(dl, {
    (ggplot(samples.l$tau %>% filter(iter == best.sample),
                    aes(x=tau, y=obstime.orig, color=capture),
                    environment=environment())
        + geom_point()
        # + scale_x_continuous(name="pseudotime")
        # + scale_y_continuous(name="capture time")
        # + scale_color_discrete(name="low-resolution\ngroup")
        + scale_y_continuous(name="true capture time")
        + scale_x_continuous(name="pseudotime")
        + scale_color_discrete(name="model\ncapture\ntime")
    )
})
print(plots$pseudo)
do.call(ggsave,
        c(list('Windram-pseudotime-vs-obfuscated.pdf', plots$pseudo + bioinf.config),
          bioinf.sizes))
if ('sample' == inf.method) {
    # Save convergence as well
    ggsave('Windram-Rhat.png', plot(dl, type="Rhat"),
        width=slide.fig.width, height=slide.fig.width / html5$ratio,
        dpi=300, units="in")
}
# Save profiles as well
plots$profiles <- plot(dl, type="profiles", genes=dl$genes.high.psi)
plots$profiles$layers[[3]]$aes_params$size <- 0.8
do.call(ggsave,
        c(list('Windram-profiles.pdf', plots$profiles + bioinf.config),
          suppl.sizes))
# Save tau posterior
plots$tau.posterior <- (
    ggplot(dl$samples.l$tau, aes(x=capture.orig, y=tau, color=capture))
    + geom_boxplot(outlier.size=.3, lwd=.15)
    # + theme(axis.text.x = element_text(angle = 90, hjust = 1))
    # + theme_gray(base_size=24)
    + scale_x_discrete(name="true capture time")
    + scale_y_continuous(name="pseudotime")
    + scale_color_discrete(name="model\ncapture\ntime")
    + coord_flip())
do.call(ggsave,
        c(list('Windram-tau-posterior.pdf', plots$tau.posterior + bioinf.config),
          bioinf.sizes))

# Save best tau
plots$tau.best <- plot(dl, type="pseudotime")
ggsave('Windram-tau-best.pdf', plots$tau.best + plos.theme,
       width=text.width, height=text.width, units="in")
ggsave('Windram-tau-best.png', plots$tau.best + plos.theme,
       width=slide.fig.width, height=slide.fig.width / html5$ratio,
       dpi=300, units="in")
```
Look at the expected correlation between the obfuscated capture time with the
pseudotime in the full posterior.
```{r}
posterior.cor <- (
    dl$samples.l$tau
    %>% group_by(iter)
    %>% dplyr::summarise(pseudotime.capture.cor=cor(tau, obstime.orig,
                                             method="spearman"))
)
posterior.cor.mean <- mean(posterior.cor$pseudotime.capture.cor)
posterior.cor.mean
posterior.cor.best <- filter(posterior.cor,
                             dl$best.sample == iter)$pseudotime.capture.cor
posterior.cor.best
monocle.cor <- 0.9269565
plots$post.cor <- (ggplot(posterior.cor, aes(x=pseudotime.capture.cor))
    + geom_histogram()
    + geom_vline(xintercept=monocle.cor, linetype='dotted')
    + geom_vline(xintercept=posterior.cor.best, linetype='dashed')
    + scale_x_continuous(name="correlation"))
print(plots$post.cor)
do.call(ggsave,
        c(list('Windram-posterior-cor.pdf', plot=plots$post.cor + bioinf.config),
          bioinf.sizes))
```
The correlation between the obfuscated capture time with the pseudotime of
the best sample.
```{r}
with(dl$samples.l$tau %>% filter(iter == dl$best.sample),
     cor(tau, obstime.orig, method="spearman"))
```

## Roughness permutation test
Permutation test for roughness.
```{r roughnesses}
dl <- roughness.test(dl, held.out.expr)
print(dl$roughness.test)
monocle.roughness <- 0.85
plots$roughnesses <- plot(dl, type="roughnesses") +
    geom_vline(xintercept=monocle.roughness, linetype='dotted')
do.call(ggsave,
        c(list('Windram-roughnesses.pdf', plots$roughnesses + bioinf.config),
          suppl.sizes))
print(plots$roughnesses)
save.output <- getOption('Windram.save.output', TRUE)
```
```{r saveDL, echo=FALSE, eval=save.output}
# Save DeLorean object without fit component
saveRDS({dl2 <- dl; dl2$fit <- NULL; dl2}, "Data/Windram.rds")
# Save plots
saveRDS(plots, "Data/Windram-plots.rds")
```
```{r loadDL, echo=FALSE, eval=FALSE}
# Not executed
# Load DeLorean object without fit component
devtools::load_all('../..')
library(DeLorean)
library(dplyr)
library(ggplot2)
dl <- readRDS("Data/Windram.rds")
plots <- readRDS("Data/Windram-plots.rds")
```

# Examine data and estimated tau using PCA
```{r PCA}
pca <- prcomp(t(dl$expr), center=TRUE, scale.=FALSE)
pca <- prcomp(t(dl$expr + dl$cell.sizes$S.hat), center=TRUE, scale.=FALSE)
# print(pca)
# summary(pca)
png('Windram-PCA-var.png')
plot(pca, type = "l")
dev.off()
pca.l <- reshape2::melt(pca$x, varnames=c("cell", "PC"), value.name="x")
# sample_n(pca.l, 10)
tau.best <- dl$samples.l$tau %>% filter(iter==dl$best.sample)
pca.df <-pca.l %>% reshape2::dcast(cell ~ PC) %>% mutate(cell=factor(cell, levels=cell.levels(dl))) %>% left_join(tau.best)
pca.plot <- ggplot(pca.df,
       aes(x=PC1, y=PC2, label=cell, color=tau, shape=capture)) +
    #geom_text() +
    geom_point() +
    scale_size_continuous(breaks=c(1,2)) +
    scale_colour_gradient(low="red", high="blue")
print(pca.plot)
ggsave('Windram-PCA.png', pca.plot)
#
# Check which tau has best log prob
extracted <- rstan::extract(dl$fit)
lapply(extracted, dim)
pars <- with(extracted, list(
    tau=tau[dl$best.sample,],
    psi=psi[dl$best.sample,],
    omega=omega[dl$best.sample,]))
pars$tauoffset = pars$tau - dl$cell.map$obstime
if (dl$opts$model.estimates.cell.sizes) {
    pars$S <- extracted$S[dl$best.sample,]
}
best.tau <- tau.for.sample(dl)
desired.tau <- sort(best.tau)
fit <- make.fit.valid(dl)
pars$tau <- best.tau
print(rstan::log_prob(fit, rstan::unconstrain_pars(fit, pars), adjust_transform=FALSE))
pars$tau <- desired.tau
print(rstan::log_prob(fit, rstan::unconstrain_pars(fit, pars), adjust_transform=FALSE))
```

R version and packages used:
```{r Rversion}
date()
sessionInfo()
```
