---
title: Analysis of Windram et al. Arabidopsis time series with out of sample predictions
author: John Reid
bibliography: DeLorean.bib
output:
  html_document:
    toc: true
    fig_caption: true
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{DeLorean analysis of Arabidopsis time series}
-->

```{r build, echo=FALSE, eval=FALSE}
devtools::load_all('../..')
# fit.model <- FALSE
fit.model <- TRUE
rmarkdown::render('Windram-PredictNew.Rmd')
```


```{r config, echo=FALSE, message=FALSE}
library(knitr)
library(knitcitations)
library(rmarkdown)
#
# knitr options
#
opts_chunk$set(
    fig.path = 'figures/Windram-',
    stop_on_error = TRUE,
    fig.width = 12.5,
    fig.height = 8)
#
# Citations
#
cleanbib()
cite_options(
    # hyperlink = 'to.doc',
    hyperlink = TRUE,
    # style = 'html',
    # citation_format = 'text',
    citation_format = "pandoc",
    cite.style = "numeric",
    check.entries = TRUE)
    # hyperlink = TRUE)
bib <- read.bibtex("DeLorean.bib")
if (file.exists("config.R")) {
    source("config.R")
}
source(system.file("scripts/shared.R", package="DeLorean"))
```

```{r loadLibs, echo=FALSE, message=FALSE}
# suppressMessages(loadfonts())
library(DeLorean)
#
# Stylesheet
#
options(markdown.HTML.stylesheet = system.file(file.path('Rmd', 'foghorn.css'),
                                               package="DeLorean"))
font.family <- "Verdana"
font.theme <- theme_update(text=element_text(family=font.family))
theme_set(font.theme)
```

`r date()`

`r citet(bib[["windram_arabidopsis_2012"]])` assayed leaves at 24 time points
in 2 conditions.


# Data

Windram et al.'s data is available in the `DeLorean` R package.
```{r loadLib}
library(DeLorean)
library(dplyr)
library(ggplot2)
data(WindramDeLorean)
seed <- getOption("Windram.seed", 1)
set.seed(seed)
```


## Obfuscate time points

Reduce resolution of observed capture time points.
```{r eval=fit.model}
group.size <- 12
windram.cell.meta$obstime.orig <- windram.cell.meta$obstime
windram.cell.meta$capture.orig <- windram.cell.meta$capture
windram.cell.meta$obstime <- (
    floor((windram.cell.meta$obstime-1) / group.size) * group.size
    + group.size / 2)
windram.cell.meta$capture <- (
    factor(as.character(windram.cell.meta$obstime),
           ordered=TRUE,
           levels=unique(as.character(windram.cell.meta$obstime))))
```


## Hold out samples
Give each sample an index, we will later use these indices to hold out samples. We will attempt to fit the held out
samples into the time course after fitting the DeLorean model.
```{r holdOut}
windram.cell.meta <-
  windram.cell.meta %>%
  group_by(condition, capture) %>%
  mutate(capture_idx = sample(n())) %>%
  ungroup()
```

Just consider the *Botrytis* cells and four samples per capture time.
```{r eval=fit.model}
dl <- de.lorean(
    windram.expr,
    windram.gene.meta,
    windram.cell.meta)
botrytis.cells <- dl$cell.meta %>% filter(condition == "Botrytis", capture_idx < 5)
dl <- filter_cells(dl, cells=botrytis.cells$cell)
```

Work out which samples are held out.
```{r calcHeldOut}
held.out.cells <-
  windram.cell.meta %>%
  filter('Botrytis' == condition, ! cell %in% dl$cell.map$cell)
```


# Estimate hyperparameters
Examine data for empirical Bayes estimation of hyperparameters.
```{r empiricalBayes, eval=fit.model}
model.name <- getOption('Windram.model', 'exact')
dl <- estimate.hyper(dl, sigma.tau=group.size/4, model.name=model.name, adjust.cell.sizes=FALSE)
if (! dl$opts$model.estimates.cell.sizes) {
    dl <- adjust.by.cell.sizes(dl)
}
fits.omega <- test.fit(dl$gene.var$omega.hat)
print(fits.omega$gp)
fits.psi <- test.fit(dl$gene.var$psi.hat)
print(fits.psi$gp)
```
```{r child=report.file("data")}
```
```{r child=report.file("hyper-parameters")}
```


# Choose genes
Choose a few (at most 100) genes to fit the model.
```{r filterGenes, eval=fit.model}
max.genes <- min(getOption("Windram.max.genes", 100))
dl <- filter_genes(dl, number=max.genes)
```
Choose some held out genes to validate the model with.
```{r heldOutGenes}
held.out.genes <- with(dl, gene.var
                           %>% left_join(gene.meta)
                           %>% filter(! gene %in% rownames(dl$expr))
                           %>% arrange(-psi.hat/omega.hat)
                           %>% head(getOption('Windram.held.out', 100)))
held.out.expr <- windram.expr[as.character(held.out.genes$gene),
                              colnames(dl$expr)]
```


# Fit model
```{r options}
num.inits <- getOption("Windram.num.inits", default.num.cores())
inf.method <- getOption("Windram.method", "vb")
```

Define and compile the model, find the best initialisation, and fit the model.
The model name is "`r model.name`", the inference method is "`r inf.method`"
with `r num.inits` initialisations, the seed is `r seed` and the data have
`r dim(dl)[1]` genes and `r dim(dl)[2]` cells.

```{r fitModel, eval=fit.model}
dl <- prepare.for.stan(dl, num.inducing.tau = 30)
dl <- compile.model(dl)
dl <- find.good.ordering(dl, seriation.find.orderings)
plot(dl, type='orderings')
dl <- pseudotimes.from.orderings(dl, num.to.keep=num.inits)
system.time(dl <- fit.model(dl, method = inf.method, num.inits = num.inits))
plots <- list()
if ("vb" == inf.method) {
    plots$pair.plot <- pseudotimes.pair.plot(dl)
}
```


```{r examConv, eval=("sample" == inf.method)}
dl <- examine.convergence(dl)
```
```{r child=report.file("convergence"), eval=("sample" == inf.method)}
```


# Analyse posterior
Examine posterior.
```{r posterior}
dl <- process.posterior(dl)
dl <- analyse.noise.levels(dl)
```
```{r child=report.file("posterior")}
```


# Profiles
Calculate expression profiles.
```{r makePredictions}
dl <- make.predictions(dl)
```
```{r child=report.file("profiles")}
```


# Fit held out samples
Fit the held out samples into pseudotime.
```{r fitHeldOutSamples}
names(dimnames(windram.expr)) <- c('gene', 'cell')
held.out.samples <-
  windram.expr[as.character(dl$gene.map$gene), held.out.cells$cell] %>%
  melt(value.name = 'expr') %>%
  mutate(
    gene = factor(gene, levels = levels(dl$gene.map$gene)),
    cell = factor(cell, levels = levels(dl$cell.map$cell))) %>%
  left_join(dl$gene.map)
# sample_n(held.out.samples, 8)
held.out.best.time <-
  held.out.samples %>%
  left_join(dl$predictions) %>%
  mutate(
    ll = dnorm(
      expr,
      mean = predictedmean + phi.hat,
      sd = sqrt(predictedvar),
      log = TRUE)) %>%
  group_by(cell, t) %>%
  summarise(ll = sum(ll)) %>%
  summarise(t = t[which.max(ll)]) %>%
  left_join(testinputs(dl)) %>%
  left_join(windram.cell.meta)
held.out.best.time
held.out.estimates <-
  held.out.samples %>%
  left_join(held.out.best.time) %>%
  left_join(dl$predictions)
# sample_n(held.out.estimates, 8)
#
# Plot original fit with new held out samples
gp <-
  profiles.plot(dl) +
  geom_point(
    data = held.out.estimates, shape = 4, size = 7,
    aes(x = tau, y = expr, colour = capture))
gp
#
# Combine original fit with held out fit
samples.combined <-
  bind_rows(
    dl$samples.l$tau %>% filter(iter == dl$best.sample) %>% select(cell, tau, capture, unobsfuscated = obstime.orig) %>% mutate(fit = 'original'),
    held.out.best.time %>% select(cell, tau, capture, unobsfuscated = obstime.orig) %>% mutate(fit = 'held out')) %>% arrange(tau)
ggplot(samples.combined, aes(x = unobsfuscated, y = tau, colour = capture, shape = fit)) + geom_point(size = 5)
```



# Examine pseudotime
Did `DeLorean` learn the obfuscated pseudotime?
```{r pseudotime}
plots$pseudo <- with(dl, {
    (ggplot(samples.l$tau %>% filter(iter == best.sample),
                    aes(x=tau, y=obstime.orig, color=capture),
                    environment=environment())
        + geom_point()
        # + scale_x_continuous(name="pseudotime")
        # + scale_y_continuous(name="capture time")
        # + scale_color_discrete(name="low-resolution\ngroup")
        + scale_y_continuous(name="true capture time")
        + scale_x_continuous(name="pseudotime")
        + scale_color_discrete(name="model\ncapture\ntime")
    )
})
print(plots$pseudo)
# Plot profiles as well
plots$profiles <- plot(dl, type="profiles", genes=dl$genes.high.psi)
plots$profiles$layers[[3]]$aes_params$size <- 0.8
# Plot tau posterior
plots$tau.posterior <- (
    ggplot(dl$samples.l$tau, aes(x=capture.orig, y=tau, color=capture))
    + geom_boxplot(outlier.size=.3, lwd=.15)
    # + theme(axis.text.x = element_text(angle = 90, hjust = 1))
    # + theme_gray(base_size=24)
    + scale_x_discrete(name="true capture time")
    + scale_y_continuous(name="pseudotime")
    + scale_color_discrete(name="model\ncapture\ntime")
    + coord_flip())
# Plot best tau
plots$tau.best <- plot(dl, type="pseudotime")
```
Look at the expected correlation between the obfuscated capture time with the
pseudotime in the full posterior.
```{r}
posterior.cor <- (
    dl$samples.l$tau
    %>% group_by(iter)
    %>% dplyr::summarise(pseudotime.capture.cor=cor(tau, obstime.orig,
                                             method="spearman"))
)
posterior.cor.mean <- mean(posterior.cor$pseudotime.capture.cor)
posterior.cor.mean
posterior.cor.best <- filter(posterior.cor,
                             dl$best.sample == iter)$pseudotime.capture.cor
posterior.cor.best
monocle.cor <- 0.9269565
plots$post.cor <- (ggplot(posterior.cor, aes(x=pseudotime.capture.cor))
    + geom_histogram()
    + geom_vline(xintercept=monocle.cor, linetype='dotted')
    + geom_vline(xintercept=posterior.cor.best, linetype='dashed')
    + scale_x_continuous(name="correlation"))
print(plots$post.cor)
```
The correlation between the obfuscated capture time with the pseudotime of
the best sample.
```{r}
with(dl$samples.l$tau %>% filter(iter == dl$best.sample),
     cor(tau, obstime.orig, method="spearman"))
```

R version and packages used:
```{r Rversion}
date()
devtools::session_info()
```
